Quinolines interfere with heme-mediated activation of artemisinins

Version 1

Abstract

Artemisinin-based combination therapies (ACTs) remain the mainstay of treatment for Plasmodium falciparum malaria, despite reports of ACT treatment failure. ACTs consist of an artemisinin and a longer-lived partner drug, which is often a quinoline. Given that heme is central to the mechanism of action of artemisinins and some quinolines, we hypothesized that these antimalarials would exhibit strong drug-drug interactions. Previous studies using standard 48 h or 72 h assays identified additive to mildly antagonistic interactions between artemisinins and quinolines. Here, we sought to re-evaluate these interactions using a pulsing assay that better mimics the short in vivo half-life of artemisinins. We found that chloroquine (CQ), piperaquine (PPQ), and amodiaquine substantially antagonize dihydroartemisinin (DHA), the active metabolite of artemisinins. CQ-DHA antagonism was notably exacerbated in CQ-resistant parasites, resulting in a superantagonistic phenotype in isobolograms. Further, we found that CQ co-treatment conferred artemisinin resistance to Kelch 13 wild type parasites in the ring stage survival assay. Using a small molecule probe to measure chemically reactive heme in live parasites, we determined that quinolines block artemisinin activation by rendering cytosolic heme inert. Finally, we probed beyond traditional ACTs, evaluating interactions of the proposed triple ACT, DHA-PPQ-Mefloquine, as well as OZ439-quinoline combinations, which were all found to be antagonistic. Collectively, these data raise concerns for the clinical use of peroxide-quinoline combination therapies.

Introduction

Malaria was responsible for an estimated 263 million cases and 597,000 deaths in 2023, the majority of which were caused by Plasmodium falciparum 1. The World Health Organization recommends artemisinin-based combination therapies (ACTs) for the treatment of P. falciparum malaria. ACTs are composed of a short-lived artemisinin derivative paired with a longer-lived partner drug. Depending on geographical resistance profiles, one of six combinations is recommended: Dihydroartemisinin (DHA) + Piperaquine (PPQ), Artesunate (AS) + Mefloquine (MFQ), AS + Amodiaquine (ADQ), AS + Pyronaridine (PYR), Artemether (AM) + Lumefantrine (LF), or AS + Sulfadoxine-pyrimethamine (SP). Three of these six partner drugs, PPQ, MFQ, and ADQ, possess quinoline groups. Importantly, heme is central to the mechanism of action of both artemisinins and quinolines 2,3.

Asexual blood stage parasites are adept at thriving in an incredibly heme rich environment. Parasites import up to 80% of host cell hemoglobin 4, which is transported to a specialized lysosome-like organelle, termed the digestive vacuole. Here, hemoglobin is proteolytically processed into globin peptides and redox active heme (Fe2+-ferroprotoporphyrin IX), which is oxidized to hemin (Fe3+-ferroprotoporphyrin IX) 5. The majority of hemin is retained within the digestive vacuole where it is detoxified via sequestration into inert hemozoin crystals 6–8. Hemin concentrations approach 0.5 M within the digestive vacuole 9. Blood stage parasites also maintain a relatively high cytosolic heme concentration of approximately 1.6 μM 10. While the source of this labile heme pool has not been experimentally determined, it is thought to originate from hemoglobin digestion.

The heme-rich environment of asexual blood stages has long been exploited for antimalarial intervention. The 4-aminoquinolines chloroquine (CQ) and PPQ are thought to act primarily in the parasite digestive vacuole, where they inhibit hemozoin formation by binding heme/hemin and the growing face of hemozoin 11–16. It has been proposed that disruption of heme homeostasis and/or accumulation of toxic heme-quinoline complexes may contribute to parasite killing 17–19. Resistance to 4-aminoquinolines is mediated by a collection of mutations in the chloroquine resistance transporter (PfCRT), which is located on the parasite digestive vacuole membrane 20. Mutations in PfCRT are thought to mediate resistance by enabling drug efflux out of the digestive vacuole 2,21–24.

Heme is also central to the mechanism of action of artemisinins. Artemisinins contain a peroxide bridge that must be cleaved by heme for antimalarial activity 25–29. Following activation, artemisinins non-specifically alkylate adjacent molecules, causing widespread cellular damage to the parasite 30–32. Clinically, artemisinin resistance is primarily associated with mutations in Kelch 13 (K13) 33–35, which are thought to reduce parasite hemoglobin uptake and digestion 36–38. Consequently, less heme is available to activate artemisinins.

ACTs remain the mainstay treatment for falciparum malaria. Widespread resistance to artemisinins in Southeast Asia and recent emergence of artemisinin resistance in Africa have posed great concern 1,34,39–42. Reduced artemisinin efficacy means that the partner drugs are responsible for clearing a larger parasite biomass, increasing the likelihood for resistance to emerge to partner drugs. Indeed, high treatment failure rates have been reported for AS-MFQ, DHA-PPQ, and AS-ADQ in Southeast Asia 43–51.

Given that heme is central to the mechanism of action of both quinolines and artemisinins, we hypothesized that these antimalarials would exhibit strong drug-drug interactions. Previously, several groups have assessed quinoline-artemisinin interactions using standard 48 or 72 h assays and found that combinations were additive in CQ-sensitive parasites and additive to slightly antagonistic in CQ-resistant parasites 52–55. However, standard 48 h or 72 h assays in which parasites are exposed to compounds for the entire assay duration do not well represent clinical drug exposure conditions and can mask not only resistance phenotypes, but also drug-drug interaction phenotypes 56–59. We sought to re-evaluate quinoline-artemisinin interactions in multidrug-sensitive, CQ-resistant, artemisinin-resistant, and PPQ-resistant parasites using pulsing assays that better represent artemisinin kinetics. Our work identifies that CQ, PPQ, and ADQ antagonize dihydroartemisinin (DHA), the active metabolite of clinically used artemisinins, by rendering heme chemically inert and preventing DHA activation. In addition, we explore drug-drug interactions beyond traditional ACTs, investigating interactions of putative triple ACTs as well as ozonide-quinoline combinations that failed in clinical trials. Overall, the data presented here highlight the importance of detailed investigation of drug-drug interactions, particularly in drug-resistant parasites.

Results

CQ and DHA are superantagonistic in CQ-resistant parasites

We began by evaluating CQ-DHA interactions in 3D7, a CQ-sensitive parasite, and Dd2, a CQ-resistant parasite (Table 1), using trophozoite-stage isobologram assays. Synchronized trophozoites were treated with fixed ratios of CQ and DHA for 4 h, drugs were washed off, and then parasitemia was assessed 72 h later. Given the short half-life of artemisinins 57, we believe that this pulsing format allows us to better assess the effect of ACT partner drugs on DHA activity compared to standard 48 or 72 h assays. Fractional IC50 (FIC50) values were calculated from fixed ratios and then plotted on an XY-graph to generate isobolograms. Isobolograms can be interpreted based on isobole shape and mean ΣFIC50 values as follows: points along the dotted line with a mean ΣFIC50 value close to 1 indicate additivity, points above the dotted line in a concave curve with a mean ΣFIC50 value >1.25 indicate antagonism, and points below the dotted line in a convex curve with a mean ΣFIC50 value <0.75 indicate synergy.

We observed strong antagonism for CQ-DHA in 3D7 parasites (Figure 1a), with a mean ΣFIC50 value of 1.68. Antagonism was notably exacerbated in Dd2 parasites (Figure 1b), which had a mean ΣFIC50 value of 2.44. The magnitude of this value far surpasses conventional antagonism, and we termed this phenomenon “superantagonism”. Perhaps even more unexpected was the atypical shape of this isobole (Figure 1b), where points extended upward and peak ΣFIC50 values exceeded 4. In contrast, peak ΣFIC50 values for 3D7 reached only 2. The traditional isobole shape for 3D7 indicates a reciprocal relationship between CQ and DHA, whereas the atypical isobole shape of Dd2 suggests a one-sided relationship, whereby CQ is blocking DHA activity.

CQ and DHA are superantagonistic in CQ-resistant parasites.

Shown are trophozoite-stage isobolograms for (a-e) CQ-DHA, (f-k) PPQ-DHA, and (l-q) MFQ-DHA. The dotted line on each graph represents perfect additivity. 3D7, Dd2, Dd2 PfCRT3D7, Dd2 PfCRTDd2, Dd2 K13R539T, and Dd2 PfCRTM343L parasites are indicated in orange, grey, pink, blue, purple, and teal, respectively. Data from three independent replicates are indicated by different shading. Figure supplement 1. CQ, DHA,sity of different parasites.

3D7 and Dd2 parasites differ genetically at a number of drug resistance markers, including PfCRT which is responsible for mediating CQ resistance. To determine if PfCRT genotype was responsible for differences in CQ-DHA interactions between these parasite lines, we evaluated interactions in two Dd2 parasites that are isogenic except at PfCRT 60,61: Dd2 PfCRT3D7, which is edited to have the 3D7 PfCRT genotype, and Dd2 PfCRTDd2, which maintains the Dd2 PfCRT genotype and was used as a cloning control (Table 1). Notably, superantagonism was lost in Dd2 PfCRT3D7 parasites, which phenocopied 3D7 parasites, and had a mean ΣFIC50 value of 1.85 and a conventional isobole shape (Figure 1c). In contrast, Dd2 PfCRTDd2 parasites displayed one-sided superantagonism with a mean ΣFIC50 value of 2.73 and peak ΣFIC50 values > 4.5 (Figure 1d). These data suggests that PfCRT genotype is the main genetic determinant that dictates CQ-DHA interactions.

Both 3D7 and Dd2 parasites are artemisinin sensitive, harboring a wildtype K13. To determine if resistance to artemisinins alters CQ-DHA interactions, we next evaluated CQ-DHA combinations in Dd2 parasites that are edited to have a K13R539T mutation 35 (Table 1). Dd2 K13R539T parasites, which also harbor a Dd2 PfCRT genotype, displayed one-sided superantagonistic isobole that was reminiscent of both Dd2 and Dd2 PfCRTDd2 parasites (Figure 1e). This suggests that, unlike PfCRT genotype, K13 genotype appeared to have no impact on drug-drug interactions at the trophozoite stage.

PPQ and DHA are antagonistic

Next, we sought to evaluate drug-drug interactions of DHA and the ACT partner drugs PPQ and MFQ. PPQ is believed to have a similar mechanism of action as CQ and both antimalarials are 4-aminoquinolines. Accordingly, we hypothesized that PPQ and DHA would also be antagonistic. In 3D7, Dd2, Dd2 PfCRT3D7, Dd2 PfCRTDd2, and Dd2 K13R539T trophozoites, we observed conventional antagonism for PPQ-DHA (Figure 1f-j) with mean ΣFIC50 values ranging from 1.55 to 1.71. To determine if, similar to CQ, PPQ-DHA interactions would be exacerbated in PPQ-resistant parasites, we also evaluated Dd2 PfCRTM343L, which harbors the Dd2 PfCRT plus an additional mutation at M343L 61 (Table 1). The M343L mutation had no impact on PPQ-DHA interactions and Dd2 PfCRTM343L parasites displayed conventional antagonism with a mean ΣFIC50 value of 1.58 (Figure 1k). Note that while the PPQ IC50 of Dd2 PfCRTM343L is 2.5-fold higher than Dd2 (65 nM versus 25 nM; Figure 1-figure supplement 1), this shift is substantially smaller than the >1000-fold CQ IC50 shift that we and others 59 have observed in CQ-sensitive versus CQ-resistant parasites using the pulsing assay format (Figure 1-figure supplement 1).

Next, we assessed interactions between MFQ and DHA (Figure 1l-q). Though MFQ contains a quinoline group, it is chemically classified as an aryl amino alcohol and is believed to have a heme-independent mechanism of action 62. Consistent with this, we observed additivity for MFQ-DHA in all six parasites tested with ΣFIC50 values ranging from 0.94 to 1.20. Overall, these data indicate that CQ and PPQ are antagonistic with DHA, while MFQ is additive.

CQ combination treatment confers artemisinin resistance to K13WT parasites in ring-stage survival assays

Given our observation that CQ and PPQ antagonize DHA in trophozoite stages, we next sought to determine if these quinolines would promote survival in early ring stages where artemisinin resistance is classically observed. We performed DHA dose-response assays on early ring stage parasites to determine both ring-stage survival assay (RSA) values and IC50 values (Figure 2a and Figure 2-figure supplement 1). The RSA is used to delineate artemisinin-sensitive versus resistant parasites, where survival > 1% corresponds to resistance and < 1% indicates sensitivity 57. Though DHA IC50 values cannot reliably differentiate artemisinin sensitive vs resistant parasites 57, they can provide important insight into different factors that contribute to artemisinin susceptibility 63.

CQ co-treatment confers artemisinin resistance in K13WT parasites.

(a) DHA dose response assays were performed on 0-3 hours post invasion (hpi) ring stage parasites. Two-fold serial dilutions of DHA were prepared in 96-well plates (DHA concentration represented by different shades of purple) starting at 1.4 μM so that both IC50 values and RSA survival values at 700 nM DHA could be determined. To evaluate quinoline-DHA interactions in early ring stages, DHA titrations were prepared with fixed concentrations of 10 μM CQ or 200 nM PPQ. (b-d) Mean RSA survival values ± SEM (top panel) and mean DHA IC50 values ± SEM (bottom panel) are shown for (b) Dd2 PfCRT3D7, (c) Dd2, and (d) Dd2 K13R539T parasites. For the PPQ-resistant parasites (e) Dd2 PfCRTM343L and (f) MRA-1284, additional fixed PPQ concentrations of 500 nM and 10 μM were also tested. For b bottom panel, a student’s t-test was used to assess statistical differences between DHA alone versus DHA + PPQ. For all other graphs, statistical differences between DHA alone (-) and DHA + quinolines was assessed using a one-way ANOVA with a Dunnett’s test for multiple comparisons. ****p < 0.0001; ***p < 0.001; *p < 0.05; ns = not significant. Figure supplement 1. Early ring stage dose-response curves.

In the absence of quinolines, the artemisinin sensitive K13WT parasites, Dd2 PfCRT3D7 and Dd2, had RSA survival values ≤ 1%, while Dd2 K13R539T had an RSA survival of 12.63% (Figure 2b-d top panel). Co-treatment with 10 μM CQ, a pharmacologically relevant concentration 64,65, significantly increased RSA survival, conferring artemisinin resistance to both Dd2 PfCRT3D7 and Dd2 parasites with mean RSA values of 2.66% and 4.49%, respectively (Figure 2b and c top panel). RSA survival values were also increased in Dd2 K13R539T parasites to 18.39%, although this shift was not statistically significant (Figure 2d top panel). When assessing DHA sensitivity by ring-stage IC50 values, CQ co-treatment also had a substantial rescuing effect in both Dd2 and Dd2 K13R539T parasites, resulting in a 4-fold increase in DHA IC50 (Figure 2c and d bottom panel). Note that while we were unable to determine a DHA IC50 value for CQ-sensitive Dd2 PfCRT3D7 parasites in the presence of 10 μM CQ, the dose-response curve was suggestive of antagonism (Figure 2-figure supplement 1a).

To determine if PPQ would have a similar rescuing effect, we performed these same assays in the presence of 200 nM PPQ. This concentration was chosen because it is sublethal for early rings in this short 3 h pulse, pharmacologically relevant 66,67, and mimics the PPQ survival assay (PSA) that is used to delineate PPQ resistance 56. Co-treatment with 200 nM PPQ did not have a rescuing effect for Dd2 PfCRT3D7, Dd2, or Dd2 K13R539T parasites and resulted in mean RSA and IC50 values that were either similar to or lower than the DHA-only control (Figure 2b-d). Notably, co-treatment with 200 nM PPQ did not shift the DHA dose response curves of these parasites (Figure 2-figure supplement 1a-c), which is consistent with a conventional antagonistic relationship.

We were curious if higher concentrations of PPQ might have a rescuing effect, particularly for PPQ-resistant parasites that are capable of surviving a higher PPQ dose. To this end, we assayed two PPQ-resistant parasites: Dd2 PfCRTM343L and MRA-1284 (IPC_6261), which is an artemisinin-resistant and PPQ-resistant clinical isolate that was collected in 2012 from Cambodia (Table 1). Dd2 PfCRTM343L and MRA-1284 have PPQ survival values (PSA) of 10% 61 and 40% 68, respectively. While co-treatment with 10 μM CQ had a rescuing effect for Dd2 PfCRTM343L as assessed by RSA and IC50 values, neither 200 nM PPQ, 500 nM PPQ, or 10 μM PPQ co-treatment rescued parasite survival (Figure 2e and Figure 2-figure supplement 1d). Similarly, 10 μM CQ co-treatment decreased MRA-1284 sensitivity to DHA as assessed by RSA and early ring IC50 values, while co-treatment with 200 nM PPQ or 500 nM PPQ had no effect on DHA sensitivity (Figure 2f). Interestingly, while co-treatment with 10 μM PPQ did not alter RSA survival of MRA-1284, it did shift the dose-response curve to the left (Figure 2-figure supplement 1e). However, since 10 μM PPQ inhibited parasite growth approximately 60% on its own, we were unable to determine a IC50 value.

CQ and PPQ antagonize DHA activation by rendering heme inert

In total, we profiled three different quinoline-DHA interactions, which are summarized in Figure 3a. In solution, quinolines and heme form heme-quinoline complexes, wherein heme is present as a µ-oxo dimer 69–71. Though the exact stoichiometry of this interaction for each quinoline remains debated 69–73, formation of such a complex could render quinoline-bound heme chemically inert (i.e. unable to activate artemisinins).

CQ and PPQ antagonize DHA activation by “inactivating” heme.

(a) Shown are mean ΣFIC50 values calculated from the trophozoite stage isobolograms in Figure 1. Values close to 1 (pink) indicate additivity. Values ranging from 1.25 to 2 (light blue) indicate classical antagonism. Values > 2.25 (dark turquoise) indicate superantagonism. (b top panel) Heme (Fe2+-FPIX) cleaves the N-O bond of H-FluNox, mediating probe fluorescence. (b bottom panel) Heme cleaves the peroxide bridge of DHA, which generates a carbon centered radical that can alkylate proximal biomolecules. (c) H-FluNox was incubated with increasing concentrations of heme in the presence or absence of 10 μM CQ, 150 nM PPQ, 10 μM PPQ, 150 nM MFQ, or 10 μM MFQ. Shown are mean relative fluorescence units (RFU) ± SD from at least three independent replicates performed in technical triplicate. (d and e) Dd2 trophozoites were treated with 10 μM CQ, 150 nM PPQ, 150 nM MFQ, or mock treated for 5.5 h. (f) Dd2 PfCRTDd2 and Dd2 PfCRT3D7 trophozoites were mock treated or treated with 150 nM CQ or 10 μM CQ for 5.5 h. (d-f) Parasites were then incubated with 10 μM Ac-H-FluNox (the cell permeable analog of H-FluNox) to visualize and quantify “active” heme. Shown are (d) representative images and (e and f) mean fluorescence intensity ± SD of at least 70 parasites from at least three independent drug treatments. Statistical significance was assessed using a one-way ANOVA with a Turkey’s test for multiple comparisons. (g and h) Dd2 trophozoites were treated with 10 μM CQ, 25 nM DHA, 10 μM CQ + 25 nM DHA for 6 h. Parasite lysates were subjected to western blot and probed with anti-K48-linked ubiquitin (K48-Ub) antibodies or anti-plasmepsin V (PMV) antibodies. Shown is a (g) representative blot and (h) quantification of relative K48-Ub intensity ± SEM from three independent replicates. Statistical significance between mock treated and antimalarial-treated parasites was determined using a one-way ANOVA with a Turkey’s test for multiple comparisons.****p < 0.0001; ***p < 0.001; **p < 0.01; ns = not significant. Figure supplement 1. Effect of pH on H-FluNox and Ac-H-FluNox activity. Figure supplement 2. Equimolar comparison of CQ, PPQ, and MFQ on Ac-H-FluNox fluorescence. Figure supplement 3. Representative images of Dd2 PfCRTDd2 and Dd2 PfCRT3D7 Ac-H-FluNox fluorescence. Figure supplement 4. Additional K48-ubiquitin western blots. Table supplement 1. Mean parasite intensity for Ac-H-FluNox experiments. Source data 1. Uncropped western blots used for ImageJ quantification.

To investigate this, we utilized a recently developed small molecule probe, H-FluNox, that is highly sensitive and specific for heme over hemin, heme-bound proteins, Fe2+, and other salts and metal ions 74. H-FluNox is basally non-fluorescent, but upon cleavage of a N-O bond by the Fe2+ of heme, H-FluNox fluoresces 74 (Figure 3b top panel). Importantly, this reaction is similar to artemisinin activation, which requires cleavage of artemisinin’s peroxide bond by the Fe2+ of heme 29 (Figure 3b bottom panel). To determine if quinoline-heme complexes would react with H-FluNox, we began by assessing H-FluNox fluorescence in solution with 10 μM CQ, which is sub-lethal for CQ-resistant parasites and pharmacologically relevant 64,65. Relative to the untreated control, 10 μM CQ inhibited fluorescence approximately 50% in the presence of 250 nM heme (Figure 3c). This suggests that CQ-heme complexes have diminished capacity to activate artemisinins. Next, we evaluated PPQ and MFQ at two different concentrations: 150 nM, which reflects a 5x trophozoite-stage IC50 concentration, and 10 μM to directly compare these quinolines with CQ. At 150 nM, both PPQ and MFQ inhibited fluorescence, albeit only 10% (Figure 3c). At 10 μM, MFQ inhibited fluorescence to a similar extent as CQ, while 10 μM PPQ only inhibited fluorescence 25% (Figure 3c). Note that 10 μM concentrations exceed maximum plasma concentrations of PPQ and MFQ, which have been reported to reach approximately 0.5-1 μM and 5 μM, respectively 66,67,75.

Next, we assessed fluorescence in live trophozoite stage parasites using the cell permeable analog of H-FluNox, Ac-H-FluNox 74. In untreated Dd2 parasites, fluorescence was predominantly observed in the parasite cytoplasm (Figure 3d), and little signal was detected in red blood cells or the parasite digestive vacuole. This suggests that a significant portion of DHA could be activated in the parasite cytoplasm. Lack of signal in the digestive vacuole could be due to issues with permeability, the specificity of Ac-H-FluNox for heme over hemin (the predominant heme species in the digestive vacuole), or sensitivity of Ac-H-FluNox to a low pH environment. In solution, H-FluNox fluorescence was decreased approximately 60% at pH 5.4 (Figure 3-figure supplement 1a), which is the estimated pH of the digestive vacuole 76. However, alkalinization of the digestive vacuole with ammonium chloride 77 did not increase digestive vacuole fluorescence signal (Figure 3-figure supplement 1b), suggesting that low pH alone cannot explain this lack of labeling.

Importantly, treatment of Dd2 trophozoites with 10 μM CQ nearly ablated Ac-H-FluNox fluorescent signal (Figure 3d and e), suggesting that CQ-heme complexes are also unable to activate artemisinins in parasites. Treatment with 150 nM PPQ significantly reduced fluorescence, though not to the same extent as 10 μM CQ (50% inhibition versus > 92% inhibition) (Figure 3d, e, and Figure 3 and 4-table supplement 1). In contrast, parasites treated with 150 nM MFQ had similar fluorescence as untreated control parasites (Figure 3d and e). These assays with H-FluNox and Ac-H-FluNox correlate well with hemozoin inhibition assays in which CQ, PPQ, and MFQ inhibit hemozoin formation in solution 70, but only CQ and PPQ inhibit hemozoin formation in parasites 7,78–80. To determine if concentration alone was responsible for the differences that we observed between CQ and the other quinolines, we also evaluated Ac-H-FluNox fluorescence following treatment with 10 μM PPQ or 10 μM MFQ. At this super-physiological and super-lethal concentration, an equimolar concentration of CQ still inhibited H-FluNox fluorescence to a greater extent than PPQ or MFQ (Figure 3-figure supplement 2 and Figure 3 and 4-table supplement 1). This suggests that something unique to the chemistry of CQ and/or the biology of the parasite is responsible for the differences that we observed.

Quinolines antagonize peroxides.

(a and b) Trophozoite-stage DHA dose-response assays were performed with Dd2 parasites in the presence or absence of 15 nM PPQ, 15 nM MFQ, 15 nM PPQ + 15 nM MFQ, or 2 nM DHA. Shown are (a) dose-response curves with mean % inhibition ± SEM and (b) mean IC50 values ± SEM from at least three independent replicates. Statistical significance was determined using a one-way ANOVA with a Turkey’s test for multiple comparisons. Trophozoite-stage isobolograms were performed on Dd2 parasites to determine drug-drug interactions of (c) PPQ-MFQ, (d) ADQ-DHA, (e) LM-DHA, (f) PPQ-OZ439, and (g) FQ-OZ439 using the following fixed ratios: 1:0, 4:1, 2:1, 1:1, 1:2, 1:4, 0:1. Shown are fractional IC50 values from at least 3 independent replicates. Independent replicates are indicated in different shading and the diagonal dotted line on each plot indicates perfect additivity. (h and i) Dd2 parasites were mock treated or treated with 10 μM CQ, 150 nM ADQ, 500 nM LM, or 150 nM FQ for 5.5 h and free heme was then labeled in live parasites with Ac-H-FluNox. Shown are (h) representative images and (i) mean fluorescence intensity ± SD of at least 90 parasites from three independent drug treatments. Statistical significance was assessed using a one-way ANOVA with a Dunnett’s test for multiple comparisons. ****p < 0.0001; ns = not significant. (j) Mean relative fluorescence of Ac-H-FluNox from drug treatments in Figure 3e and 4i was plotted against Dd2 mean ΣFIC50 values for the indicated combinations. Combinations with DHA are indicated in black. Combinations with OZ439 are indicated in blue. Figure supplement 1. Additional MFQ isobolograms. Figure supplement 2. ADQ, LM, and FQ IC50 values. Table supplement 1. Mean parasite intensity for Ac-H-FluNox experiments.

PfCRTDd2 is believed to mediate CQ resistance by pumping CQ out of the digestive vacuole 2,21–24, suggesting that cytoplasmic CQ concentrations may differ between PfCRTDd2 and PfCRT3D7 parasites treated with the same concentration of CQ. Given the substantial difference that we saw in the shape of CQ-DHA isobolograms between PfCRTDd2 and PfCRT3D7 parasites, we were curious if differences in PfCRT genotype would impact Ac-H-FluNox fluorescence. Accordingly, we treated Dd2 PfCRTDd2 and Dd2 PfCRT3D7 parasites with 150 nM CQ or 10 μM CQ and measured chemically “reactive” heme using Ac-H-FluNox. In untreated Dd2 PfCRTDd2 versus Dd2 PfCRT3D7 parasites, we observed no difference in Ac-H-FluNox localization or fluorescence intensity (Figure 3f and Figure 3-figure supplement 3). Following treatment with 150 nM CQ, fluorescence was inhibited 25% in Dd2 PfCRT3D7 and 35% in Dd2 PfCRTDd2, although this difference was not statistically significant (Figure 3f and Figure 3 and 4-table supplement 1). Curiously, while 10 μM CQ treatment reduced mean fluorescence approximately 60% in Dd2 PfCRT3D7 parasites, it was not to the extent of Dd2 PfCRTDd2 parasites (80%) (Figure 3f and Figure 3 and 4-table supplement 1). These data suggest that transport of CQ out of the digestive vacuole by PfCRTDd2 may contribute to superantagonism between CQ and DHA in CQ-resistant parasites.

CQ co-treatment rescues DHA-mediated protein damage

Following activation by heme, DHA causes widespread protein damage in the parasite 30–32,81. Accordingly, we would expect that if CQ is antagonizing DHA activation, then CQ co-treatment should rescue DHA-mediated protein damage. To test this hypothesis, we treated Dd2 trophozoites with CQ, DHA, or CQ + DHA and probed for K48-linked ubiquitin, which is a marker of protein damage 81. Consistent with what has been previously reported 82, CQ treatment alone had no effect on K48-linked ubiquitin levels, while treatment with DHA resulted in a 2-fold increase in K48-linked ubiquitin (Figure 3g, h, and Figure 3-figure supplement 4). Importantly, addition of CQ to DHA had a rescuing effect, resulting in levels of K48-linked ubiquitin that were similar to the DMSO-treated control (Figure 3g and h). Collectively, these data suggest that CQ antagonizes DHA by rendering heme chemically inert, reducing DHA activation, and consequently, reducing DHA-mediated protein damage.

PPQ and MFQ do not harm treatment efficacy, but strongly antagonize each other

With the looming threat of widespread ACT treatment failure, several ideas have been proposed to delay the spread of ACT resistance by using currently available antimalarials. Triple ACTs (TACTs), which are composed of a traditional ACT plus an additional longer-lived partner drug, have gained the most traction. Early clinical trials indicate that DHA-PPQ-MFQ and AM-LM-ADQ show good safety and efficacy against artemisinin-sensitive and resistant parasites 47,83–85. We were curious how the addition of a third antimalarial would influence drug interactions. Accordingly, we performed trophozoite-stage DHA dose response assays on Dd2 parasites in the presence of a fixed, sub-lethal concentration of PPQ, MFQ, or PPQ + MFQ. Addition of 2 nM DHA was also included as a control to represent additivity. As assessed by DHA IC50 values, addition of 15 nM PPQ did not improve parasite killing, while addition of 15 nM MFQ resulted in a 3-fold decrease in IC50 compared to the DMSO control (Figure 4a and b). These phenotypes are consistent with antagonistic and additive relationships, respectively. Notably, addition of PPQ + MFQ had no additional benefit compared to addition of MFQ alone (Figure 4 a and b).

Following the initial 3-day TACT treatment regimen, long-lived partner drugs are responsible for clearing the remaining parasite burden. Accordingly, we sought to assess drug-drug interactions of PPQ-MFQ in the absence of DHA. Previously it was reported that MFQ and PPQ were moderately antagonistic in 3D7 and K1 parasites using a standard 48 h assay, with mean ΣFIC50 values of 1.31 and 1.28, respectively 52. Using our trophozoite-stage assay, we observed reciprocal superantagonism in both PPQ-sensitive, Dd2 (Figure 4c), and PPQ-resistant Dd2 PfCRTM343L (Figure 4-figure supplement 1a), with mean ΣFIC50 values of 2.62 and 2.44, respectively. This phenotype appears to be unique to PPQ, as we observed only strong antagonism for CQ-MFQ with a mean ΣFIC50 of 1.72 (Figure 4-figure supplement 1b). Overall, these data suggest that while adding PPQ and MFQ to DHA does not harm treatment efficacy, PPQ and MFQ may not be a strategic antimalarial pairing for TACTs given their superantagonism with each other.

Quinolines antagonize peroxides

To determine if the drug-drug interactions that we evaluated were specific to a subset of antimalarials or represented a broader phenotype, we next evaluated DHA interactions with the clinically used ACT partner drugs ADQ and LM. Structurally, ADQ is classified as a 4-aminoquinoline like PPQ and CQ, while LM is classified as an aryl amino alcohol like MFQ. However, unlike MFQ, LM lacks a quinoline group. We found that in Dd2 parasites, ADQ and DHA were antagonistic (Figure 4d) with a mean ΣFIC50 of 1.61, while LM and DHA were additive (Figure 4e) with a mean ΣFIC50 of 1.02.

To determine if 4-aminoquinolines antagonize other peroxide antimalarials, we next sought to assess interactions of two different OZ439-quinoline pairs that were previously evaluated in clinical trials. Similar to artemisinins, the synthetic ozonide, OZ439, requires activation via heme-mediated cleavage of a peroxide bond. In phase II clinical trials, OZ439 showed good efficacy alone 86, but poor efficacy either in combination with PPQ 87 or the next generation quinoline, ferroquine (FQ) 88,89. To assess drug-drug interactions, we performed trophozoite-stage isobolograms with OZ439-PPQ and OZ439-FQ in Dd2 parasites. Both PPQ-OZ439 (Figure 4f) and FQ-OZ439 (Figure 4g) were antagonistic with a mean ΣFIC50 values of 1.42 and 1.52, respectively. These data suggest that antagonistic interactions between OZ439 and quinolines may, in part, have contributed to the poor efficacy of these single dose combination therapies.

Although the mechanism of action of ADQ and FQ are less well studied, both antimalarials appear to inhibit hemozoin formation. ADQ was shown to inhibit hemozoin formation in parasites using the heme fractionation assay 80. While in-parasite assays have not been performed with FQ, in vitro biochemical studies suggest that FQ may inhibit hemozoin formation even better than CQ 90. Accordingly, we were curious if ADQ and FQ treatment would “inactivate” heme in parasites. Using the Ac-H-FluNox heme probe, we observed that relative to the untreated control, treatment with 150 nM ADQ or 150 nM FQ (approximate 5x IC50 concentrations; Figure 4-figure supplement 2) reduced Ac-H-FluNox fluorescent signal approximately 50% and 55%, respectively (Figure 4h, i, and Figure 3 and 4-table supplement 1). In contrast, treatment with an equipotent concentration of LM had no significant impact on Ac-H-FluNox signal relative to the untreated control (Figure 4h and i). Among the six partner drugs that were evaluated, we observed a clear inverse correlation between Ac-H-FluNox signal and mean ΣFIC50, with an R2 value of 0.84 (Figure 4j). These data strongly suggest that antimalarials that render heme inert also antagonize artemisinin and ozonide antimalarials, presumably by blocking endoperoxide activation.

Discussion

Combination therapies are widely believed to delay the emergence of drug resistance not only for malaria, but also bacterial infections, HIV, fungal infections, and cancer. However, to maximize the benefits of combination therapies, it is essential that drugs are strategically paired. At a minimum, combinations should enhance efficacy to reduce parasite biomass and reduce the potential for drug resistance to evolve. In vitro assessment of drug-drug interactions between two antimalarials with vastly different half-lives is challenging. We chose to use a short treatment pulse to mimic artemisinin exposure and gain an understanding into how quinolines impact artemisinin activity. While drug-drug interactions are undoubtably much more complex in vivo, we believe this assay format better recapitulates clinical conditions compared to traditional 48 or 72 h assays.

We found that CQ, PPQ, and ADQ are antagonistic with DHA as assessed in early ring and trophozoite stages. Our work suggests that these antimalarials antagonize DHA activation by rendering intra-parasitic heme chemically inert. To date, it remains debated where artemisinins are activated within the parasite. There are several lines of evidences that suggest that artemisinins can be activated in the parasite cytoplasm. First, immunofluorescence studies with an anti-AS antibody 91 and live cell imaging with an artemisinin-based photoaffinity probe 92 revealed that artemisinins localize throughout the parasite cytoplasm. Second, numerous chemical proteomics studies utilizing artemisinin and ozonide-based click chemistry probes have identified a number of cytoplasmic proteins as targets of artemisinin and ozonides 30–32,93,94. Given how quickly activated artemisinins react with adjacent molecules, it is unlikely that activated artemisinins could traverse the digestive vacuole or other organelle membranes before encountering a cytoplasmic target. Finally, immunofluorescence assays with antibodies specific for the alkylation signature of the ozonide OZ277 95 revealed labeling throughout the parasite cytoplasm. Our data with Ac-H-FluNox also lend credence to this hypothesis. We observed Ac-H-FluNox fluorescence throughout the parasite cytoplasm, indicating that there is a substantial amount of heme in this location that is chemically competent to activate artemisinins. It is possible that technical limitations, such as issues with membrane permeability, limit Ac-H-FluNox fluorescence in the digestive vacuole. Nevertheless, these data collectively indicate that artemisinins can be activated in the parasite cytoplasm.

Previously, it has been reported that CQ treatment increases intracellular “free” heme 7,10,96. However, the term “free” heme is somewhat malleable. While the pyridine-heme fractionation assay 7,80, electron spectroscopic imaging (ESI) with energy-loss spectroscopy (EELS) 7 and heme biosensor 10, can be used to detect and quantify heme species, they are unable to differentiate chemically “active” heme from inert heme (i.e. quinoline bound or protein bound). In contrast, H-FluNox is highly specific for chemically “active” heme 74. One possibility to reconcile our data with previous reports is that CQ treatment increases non-hemozoin/hemoglobin heme, but that this heme is bound to CQ or other biomolecules, rendering it unable to activate peroxides. Importantly, activation of Ac-H-FluNox is very similar to that of artemisinins, providing a useful proxy to assess not only chemically active heme, but also artemisinin activation in live parasites.

CQ-DHA antagonism was notably exacerbated in CQ-resistant parasites. The concentrations used in isobolograms are chosen based on parasite IC50 values to individual compounds. While all parasites were exposed to similar DHA concentrations, CQ-resistant parasites were exposed to much higher CQ concentrations compared to their CQ-sensitive counterparts. This may explain in part why antagonism was intensified in CQ resistant parasites. However, equimolar treatment with 10 µM CQ resulted in significantly lower Ac-H-FluNox signal in Dd2 PfCRTDd2 versus Dd2 PfCRT3D7 parasites (Figure 3f). These data could suggest that while the majority of CQ is sequestered in the digestive vacuole of Dd2 PfCRT3D7 parasites, PfCRTDd2 is able to transport more CQ out of the digestive vacuole, thus allowing more CQ to bind cytosolic heme in parasites harboring a PfCRTDd2. This duality of PfCRTDd2, simultaneously enhancing parasite survival to CQ while promoting cytoplasmic heme “inactivation”, may explain the unprecedented one-sided superantagonistic isobole that we observed.

Fortunately, we did not observe superantagonism with DHA-PPQ in PPQ-resistant parasites. With a modest 3-fold shift in PPQ IC50 between Dd2 vs Dd2 PfCRTM343L parasites (25 nM vs 65 nM), it is perhaps unsurprising that no significant difference in DHA-PPQ isobolograms was observed between these two parasites. This is in contrast to the >1000-fold CQ IC50 shift that we and others 59 have observed between CQ-sensitive and CQ-resistant parasites using pulsing assays. Our studies with Ac-H-FluNox indicate that 150 nM PPQ is sufficient to render approximately 50% of available heme inert (Figure 3e). While these data suggest that PPQ blocks some DHA activation, it is likely that 150 nM PPQ is not sufficient to block all DHA activation. Since even PPQ-resistant parasites are at least partially susceptible to mid-nanomolar PPQ concentrations, the protective effect that PPQ has in neutralizing DHA is partially negated by PPQ-mediated killing. In contrast, CQ-resistant parasites are relatively unaffected by pulses with low micromolar CQ concentrations. Accordingly, CQ concentrations that are high enough to nearly block all DHA activation are insufficient to kill CQ resistant parasites.

Relative to PPQ and MFQ, CQ appears to have the greatest ability to render heme inert. We were surprised to observe that PPQ inhibited H-FluNox fluorescence only half as well as CQ in solution. This suggesting that there may be important differences in how these 4-aminoquinolines bind heme. In Dd2 parasites, CQ also inhibited Ac-H-FluNox fluorescence to a greater extent than PPQ at 10 μM concentrations. Given that the ability of CQ to “defuse” heme in parasites is dependent on PfCRT genotype, it is possible that these results may differ in a PPQ-resistant parasite. In solution, equimolar concentrations of CQ and MFQ inhibited H-FluNox fluorescence to a similar extent, while MFQ had a minimal effect on Ac-H-FluNox fluorescence at even super-pharmacological concentrations. This was somewhat surprising given that MFQ is believed to act in the parasite cytoplasm 97,98. It possible that differences in how MFQ binds to heme 70, affinity of MFQ for other biomolecules 99, and/or modification of MFQ within the parasite could contribute to this discrepancy.

In the early 2000’s, the rise of CQ resistance prompted a search for new combination therapies 100. Several clinical trials were conducted in West Africa to compare efficacy of CQ versus CQ-AS. On day 28, two studies reported similar failure rates for CQ and CQ-AS 101,102. In in a third study, cure rates were significantly higher for CQ-AS compared to CQ alone 103. However, all three studies recommended against use of CQ-AS, due to poor treatment efficacy. Notably, AS-SP and AS-LM were reported to show > 97% efficacy on West African populations during the same time period 104,105. For all currently used ACTs, treatment failure appears to be dependent on parasites acquiring resistance to both artemisinins and partner drugs. It is difficult to speculate if drug-drug interactions could have played a role in resistance evolution. Nevertheless, it remains questionable whether use of DHA-PPQ or AS-ADQ should be enthusiastically supported, given that PPQ and ADQ appear to directly block artemisinin activation.

With the absence of new antimalarials ready for clinical implementation, novel strategies with existing drugs have been proposed to overcome current problems with ACT treatment failures. One such strategy, TACTs, has gained the most traction. For traditional ACTs, the short one hour half-life of artemisinins means that long-lived partner drugs are essentially acting as monotherapies for most of the initial 3-day treatment regimen and a substantial period thereafter. In the case of artemisinin-resistant parasites, partner drugs must clear a significantly higher parasite biomass, which increases the likelihood for resistance to emerge. In theory, TACTs are designed to overcome this flaw, whereby two partner drugs with similar half-lives are used, ensuring that parasites are always exposed to at least two antimalarials 106,107. In clinical trials, DHA-PPQ-MFQ showed improved efficacy over DHA-PPQ in regions of high PPQ-resistance 47. However, some have speculated that this was solely due to application of MFQ to MFQ-sensitive parasites and that the TACT would be no more effective than DHA-MFQ 108. PPQ and MFQ were initially promoted as ideal partner drugs for TACTs as it was thought that either due to collateral sensitivity or fitness costs, dual PPQ and MFQ resistance was unlikely to occur 109,110. Use of edited isogenic lines suggests that PPQ resistance conferring mutations in PfCRT have no effect on MFQ sensitivity, while increased MDR1 copy number, which underlies MFQ-resistance, has no effect on PPQ sensitivity 78. Further, more recent genetic surveillance studies of Cambodian parasites cast doubt on the hypothesis that a DHA-PPQ-MFQ resistant parasite would be unfit to spread clinically 111. Importantly, we observed superantagonism between PPQ and MFQ in trophozoite-stage isobolograms. This is consistent with previous reports indicating that PPQ and MFQ are antagonistic as assessed with standard 48 or 72 h assays 52,55. The double antagonism between PPQ-DHA and MFQ-PPQ likely explains why we observed no improvement in efficacy between DHA-MFQ and DHA-MFQ-PPQ treatments (Figure 4a and b). With the interplay of existing PPQ-resistance, PPQ-MFQ superantagonism, and any potential parasites fitness cost, it is difficult to estimate how quickly DHA-PPQ-MFQ resistant parasites would emerge and spread. Nevertheless, these data collectively indicate that PPQ and MFQ should be scrutinized as partners for TACTs.

The synthetic ozonide, OZ439, was previously poised to replace artemisinins in combination therapies. While OZ439 is believed to have a mechanism of action similar to that of artemisinins, it has a notably improved 46-62 h half-life 86 and was shown to be effective against artemisinin-resistant parasites 112–114. While OZ439 showed good efficacy alone 86, high treatment failure as part of single-dose combination therapies ultimately resulted in loss of interest of this antimalarial 87–89. However, our data suggest that unsuitable choices in partner drugs may be partly responsible for the poor therapeutic efficacy of these combinations. Both PPQ and FQ rendered heme inert, directly blocking OZ439 activation. Biochemical studies have predicted that due to its less basic behavior, FQ might have a lower propensity to accumulate in the parasite digestive vacuole compared to CQ 90. Greater accumulation of FQ in the parasite cytoplasm compared to other quinolines might explain the ability of FQ to greatly diminish Ac-H-FluNox signal even at nanomolar concentrations. Clinically, peak levels of FQ have been reported to reach approximately 0.5 μM 115, which could significantly impact OZ439 activation. However, unlike CQ, mid-nanomolar FQ concentrations are lethal for parasites which may explain why we did not observe a super antagonistic phenotype for FQ-OZ439.

The antimalarials that we have are a precious resource, and it is only a matter of time before they fail. Our work clearly demonstrates the importance of detailed understanding of drug-drug interactions at the parasite level. While we believe the pulsing assay used in this study better mimics artemisinin pharmacokinetics compared to standard 72 h assays, it is not suitable for high throughput format. Recently, a high throughput assay was developed to evaluate drug-drug and drug-drug-drug interactions in P. falciparum 55. Using this assay, the authors were able to evaluate interactions of 16 potential TACT combinations against 13 clinical isolates 55. Similar to our findings, the authors reported that interactions can be highly dependent on parasite genetics. This high throughput assay will undoubtably be an important first step in evaluating interactions of new combinations. Nevertheless, since this assay relies on a 72 h format, additive interactions should be interpreted with caution, as this format can mask drug-drug interactions 58. Accordingly, pulsing assays like the one used in this study will be an important step for investigating drug-drug interactions, prior to in vivo studies 116.

Materials and Methods

Reagents

All reagents were purchased from Sigma-Aldrich unless otherwise indicated.

Parasite Culture

Dd2 PfCRTDd2, Dd2PfCRT3D7, Dd2 PfCRTM343L, and Dd2 K13R539T were kindly provided by Dr. David Fidock (Columbia University) 35,61. MRA-1284 was obtained through BEI Resources, NIAID, NIH: Plasmodium falciparum, Strain IPC_6261, MRA-1284, contributed by Didier Ménard. Parasites were propagated in human red blood cells obtained from St. Louis Children’s Hospital blood bank. Cultures were grown at 5% hematocrit in media containing RPMI1640 (Gibco) supplemented with 0.25% (wt/vol) Albumax (Gibco), 15 mg/L hypoxanthine, 110 mg/L sodium pyruvate, 1.19 g/L HEPES, 2.52 g/L sodium bicarbonate, 2 g/L glucose, and 10 mg/L gentamicin and were maintained in air-tight chambers filled with 5% O2/5%CO2/90%N2 at 37°C.

Trophozoite-Stage Dose-Response Assays and Isobolograms

Trophozoite-stage assays were performed as previously described with minor modifications 117. Briefly, parasites were synchronized to the trophozoites stage using two consecutive treatments with 5% sorbitol 10 h apart, followed by culturing for an additional 14 h. Trophozoite stages were seeded at 0.2% parasitemia and 1% hematocrit in round-bottom 96-well plates. Following 4 h incubation with antimalarials, infected red blood cells were washed a minimum of 5 times for plates containing OZ439, and 4 times for all other antimalarials. Plates were centrifuged, 190 μL of media was carefully removed, and 190 μL of fresh media was added. Following the last wash, cultures were transferred to a new flat-bottom 96-well plate. After 72 hr, parasitemia was assessed by flow cytometry. Cultures were resuspended and 10 μL was transferred to a new 96-well plate containing 50 μL per well of 1 x PBS with 1 x SYBR Green I and 100 nM MitoTracker Deep Red. Following incubation at 37°C for 1 h, plates were read on a BD FACSCanto with a BD HTS automated plate reader and analyzed with BD FACSDiva Software version 8. Approximately 10,000 events were analyzed per well. GraphPad version 10 was used to calculate IC50 values by nonlinear regression.

For dose response assays with fixed concentrations of partner drugs, a 96-well plate was prepared containing 2-fold serial dilutions of DHA media at 100 μL per well. A bulk intermediate culture of parasites was prepared at 2% hematocrit. Partner drugs were added to this intermediate culture at 2x concentrations. After mixing, 100 μL of the intermediate culture was added to the 96-well plate containing DHA serial dilutions, for a final hematocrit of 1% and a 1 x fixed partner drug concentration.

For isobolograms, the following fixed ratios were prepared as starting concentrations: 1:0, 4:1, 2:1, 1:1, 1:2, 1:4, 0:1. The starting concentration of each drug alone (1:0 and 0:1 ratios) represents an approximate 16x IC50 concentration. Trophozoite stages were exposed to 2-fold dilutions of these ratios. Fractional IC50 (FIC50) values were calculated as follows: FIC50 values were plotted on an XY graph. Mean ΣFIC50 values were calculated by averaging the sums of FIC50 values from 4:1, 2:1, 1:1, 1:2, and 1:4 ratios. Trophozoite-stage IC50 values for Figure 1-figure supplement 1 and Figure 4-figure supplement 2 were obtained from 1:0 or 0:1 ratios. A one-way ANOVA with a Turkey’s or Dunnett’s multiple comparison test were used to assess statistical significance for IC50 values.

RSAs and Early Ring Stage Dose Response Assays

0-3 hours post invasion (hpi) rings were obtained as previously described 118. Briefly, schizonts were enriched using MACS LD magnet columns (Miltenyi Biotec). Schizont enriched cultures were diluted to 2% hematocrit in 3-4 mL of media and incubated for 3 h with shaking. A 5% sorbitol treatment was then used to remove remaining mature parasites. Early ring stages were seeded at 1% parasitemia and 1% hematocrit in round-bottom 96-well plates. 2-fold serial dilutions were prepared starting at 1400 nM DHA, so that both IC50 values and RSA values could be obtained. After incubation for 3 h, antimalarials were washed out and parasitemia was assessed 72 hr later by flow cytometry as described in the previous section. GraphPad Prism version 10 was used to calculate IC50 values by nonlinear regression. RSA survival was obtained by dividing the parasitemia of cultures treated with 700 nM DHA by the untreated control. A one-way ANOVA with a Dunnett’s multiple comparison test was used to determine statistical significance for IC50 values and RSA survival values.

Detection of Free Heme In Vitro

H-FluNox was kindly provided by Dr. Tasuku Hirayama (Gifu Pharmaceutical University, Lab of Pharmaceutical & Medicinal Chemistry). Reactions were carried out in a buffer containing 50 mM HEPES (pH 7.4) and 100 μM reduced glutathione. Glutathione was used to convert hemin into heme. Hemin was made up in reaction buffer and 2-fold serial dilutions were performed to obtain a range of heme concentrations starting at 500 nM. H-FluNox was used at a final concentration of 500 nM. Reactions were performed in black-walled clear-bottom 96-well plates. Following incubation with H-FluNox and indicated antimalarials at room temperature for 30 min, fluorescence was read at 485/535 nm ex/em using an EnVision Multimode Plate Reader.

Detection of Free Heme In Live Parasites

Ac-H-FluNox was kindly provided by Dr. Tasuku Hirayama (Gifu Pharmaceutical University, Lab of Pharmaceutical & Medicinal Chemistry). Parasites were synchronized to the trophozoite stage as described above. Cultures were then treated with the indicated compound for the indicated time. Media was removed and parasite cultures were resuspended and incubated in 1 x PBS containing 10 μM Ac-H-FluNox for 30 min at 37°C. Ac-H-FluNox is the cell permeable analog of H-FluNox and is converted to H-FluNox in cells 74. Following incubation, cultures were washed with 1x PBS and then imaged using a 63× objective on a Zeiss Imager M2 Plus Widefield fluorescence microscope with AxioVision 4.8 software. Except for Figure 3-figure supplement 1b, a fixed exposure of 350 ms was used to capture H-FluNox fluorescence at 470/509 nm ex/em. Exposure for Figure 3-figure supplement 1b was adjusted for each condition so that DV staining could be optimally visualized. For quantification in Figure 3e, 3f, 4i, and Figure 3-figure supplement 2, at least ten fields were captured per treatment condition and then mean fluorescence intensity of parasite minus background was quantified with ImageJ. Statistical significance of ≥ 70 parasites from at least three independent drug treatments was assessed using a one-way ANOVA.

Western Blot

Parasite drug treatment, protein harvest, sample preparation, SDS-PAGE, and wet transfer were performed as previously described 82. Blots were blocked with 3% BSA in TBS-T and incubated overnight with primary antibodies used at 1:1000 in 1x TBS-T. Anti K48-linked ubiquitin was obtained from Cell Signaling Technologies (catalog number: D9D5, rabbit). Anti-PMV was obtained from 119. Blots were incubated for 1 h at room temperature with IRDye conjugated goat secondary antibodies (LICOR) at 1:10,000 dilutions. Blots were imaged using a Bio Rad imager with Image Lab Touch Software and then quantified with ImageJ. Significant differences between control and antimalarial treated parasites were determined using a one-way ANOVA with a Dunnett’s multiple comparison test.

eLife Assessment

This study is important as it demonstrates that 4-aminoquinoline antimalarials antagonize artemisinin activity under physiologically relevant conditions. Using isogenic parasite lines and a chemical probe, the authors provide mechanistic insight and compelling evidence implicating PfCRT in this antagonism. However, some weaknesses have been identified that limit full interpretation of the findings, which are based solely on in vitro assays, though the results have implications that will be of importance in optimizing future antimalarial combination strategies.

Reviewer #1 (Public review):

Summary:

This study set out to investigate potential pharmacological drug-drug interactions between the two most common antimalarial classes, the artemisinins and quinolines. There is a strong rationale for this aim, because drugs from these classes are already widely used in Artemisinin Combination Therapies (ACTs) in the clinic, and drug combinations are an important consideration in the development of new medicines. Furthermore, whilst there is ample literature proposing many diverse mechanisms of action and resistance for the artemisinins and quinolines, it is generally accepted that the mechanisms for both classes involve heme metabolism in the parasite, and that artemisinin activity is dependent on activation by reduced heme. The study was designed to measure drug-drug interactions associated with a short pulse exposure (4 h) that is reminiscent of the short duration of artemisinin exposure obtained after in vivo dosing. Clear antagonism was observed between dihydroartemisinin (DHA) and chloroquine, which became even more extensive in chloroquine-resistant parasites. Antagonism was also observed in this assay for the more clinically-relevant ACT partner drugs piperaquine and amodiaquine, but not for other ACT partners mefloquine and lumefantrine, which don't share the 4-aminoquinoline structure or mode of action. Interestingly, chloroquine induced an artemisinin resistance phenotype in the standard in vitro Ring-stage Survival Assay, whereas this effect was not apparent for piperaquine.

The authors also utilised a heme-reactive probe to demonstrate that the 4-aminoquinolines can inhibit heme-mediated activation of the probe within parasites, which suggests that the mechanism of antagonism involves the inactivation of heme, rendering it unable to activate the artemisinins. Measurement of protein ubiquitination showed reduced DHA-induced protein damage in the presence of chloroquine, which is also consistent with decreased heme-mediated activation, and/or with decreased DHA activity more generally.

Overall, the study clearly demonstrates a mechanistic antagonism between DHA and 4-aminoquinoline antimalarials in vitro. It is interesting that this combination is successfully used to treat millions of malaria cases every year, which may raise questions about the clinical relevance of this finding. However, the conclusions in this paper are supported by multiple lines of evidence, and the data are clearly and transparently presented, leaving no doubt that DHA activity is compromised by the presence of chloroquine in vitro. It is perhaps fortunate that the clinical dosing regimens of 4-aminoquinoline-based ACTs have been sufficient to maintain clinical efficacy despite the non-optimal combination. Nevertheless, optimisation of antimalarial combinations and dosing regimens is becoming more important in the current era of increasing resistance to artemisinins and 4-aminoquinolines. Therefore, these findings should be considered when proposing new treatment regimens (including Tripe-ACTs) and the assays described in this study should be performed on new drug combinations that are proposed for new or existing antimalarial medicines.

Strengths:

This manuscript is clearly written, and the data presented are clear and complete. The key conclusions are supported by multiple lines of evidence, and most findings are replicated with multiple drugs within a class, and across multiple parasite strains, thus providing more confidence in the generalisability of these findings across the 4-aminoquinoline and peroxide drug classes.

A key strength of this study was the focus on short pulse exposures to DHA (4 h in trophs and 3 h in rings), which is relevant to the in vivo exposure of artemisinins. Artemisinin resistance has had a significant impact on treatment outcomes in South-East Asia, and is now emerging in Africa, but is not detected using a 'standard' 48 or 72 h in vitro growth inhibition assay. It is only in the RSA (a short pulse of 3-6 h treatment of early ring stage parasites) that the resistance phenotype can be detected in vitro. Therefore, assays based on this short pulse exposure provide the most relevant approach to determine whether drug-drug interactions are likely to have a clinically relevant impact on DHA activity. These assays clearly showed antagonism between DHA and 4-aminoquinolines (chloroquine, piperaquine, amodiaquine, and ferroquine) in trophozoite stages. Interestingly, whilst chloroquine clearly induced an artemisinin-resistant phenotype in the RSA, piperaquine did not appear to impact the early ring stage activity of DHA, which may be fortunate considering that piperaquine is a currently recommended DHA partner drug in ACTs, whereas chloroquine is not!

The evaluation of additional drug combinations at the end of this paper is a valuable addition, which increases the potential impact of this work. The finding of antagonism between piperaquine and OZ439 in trophozoites is consistent with the general interactions observed between peroxides and 4-aminoquinolines, and it would be interesting to see whether piperaquine impacts the ring-stage activity of OZ439.

The evaluation of reactive heme in parasites using a fluorescent sensor, combined with the measurement of K48-linked ubiquitin, further supports the findings of this study, providing independent read-outs for the chloroquine-induced antagonism.

The in-depth discussion of the interpretation and implications of the results is an additional strength of this manuscript. Whilst the discussion section is rather lengthy, there are important caveats to the interpretation of some of these results, and clear relevance to the future management of malaria that require these detailed explanations.

Overall, this is a high-quality manuscript describing an important study that has implications for the selection of antimalarial combinations for new and existing malaria medicines.

Weaknesses:

This study is an in vitro study of parasite cultures, and therefore, caution should be taken when applying these findings to decisions about clinical combinations. The drug concentrations and exposure durations in these assays are intended to represent clinically relevant exposures, although it is recognised that the in vitro system is somewhat simplified and there may be additional factors that influence in vivo activity. I think this is reasonably well acknowledged in the manuscript.

It is also important to recognise that the majority of the key findings regarding antagonism are based on trophozoite-stage parasites, and one must show caution when generalising these findings to other stages or scenarios. For example, piperaquine showed clear antagonism in trophozoite stages, but not in ring stages under these assay conditions.

The key weakness in this manuscript is the over-interpretation of the mechanistic studies that implicate heme-mediated artemisinin activation as the mechanism underpinning antagonism by chloroquine. In particular, the manuscript title focuses on heme-mediated activation of artemisinins, but this study did not directly measure the activation of artemisinins. The data obtained from the activation of the fluorescent probe are generally supportive of chloroquine suppressing the heme-mediated activation of artemisinins, and I think this is the most likely explanation, but there are significant caveats that undermine this conclusion. Primarily, the inconsistency between the fluorescence profile in the chemical reactions and the cell-based assay raises questions about the accuracy of this readout. In the chemical reaction, mefloquine and chloroquine showed identical inhibition of fluorescence, whereas piperaquine had minimal impact. On the contrary, in the cell, chloroquine and piperaquine had similar impacts on fluorescence, but mefloquine had minimal impact. This inconsistency indicates that the cellular fluorescence based on this sensor does not give a simple direct readout of the reactivity of ferrous heme, and therefore, these results should be interpreted with caution. Indeed, the correlation between fluorescence and antagonism for the tested drugs is a correlation, not causation. There could be several reasons for the disconnect between the chemical and biological results, either via additional mechanisms that quench fluorescence, or the presence of biomolecules that alter the oxidation state or coordination chemistry of heme or other potential catalysts of this sensor. It is possible that another factor that influences the H-FluNox fluorescence in cells also influences the DHA activity in cells, leading to the correlation with activity. It should be noted that H-FluNox is not a chemical analogue of artemisinins. Its activation relies on Fenton-like chemistry, but with an N-O rather than O-O bond, and it possesses very different steric and electronic substituents around the reactive centre, which are known to alter reactivity to different iron sources. Despite these limitations, the authors have provided reasonable justification for the use of this probe to directly visualise heme reactivity in cells, and the results are still informative, but additional caution should be provided in the interpretation, and the results are not conclusive enough to justify the current title of the paper.

Another interesting finding that was not elaborated by the authors is the impact of chloroquine on the DHA dose-response curves from the ring stage assays. Detection of artemisinin resistance in the RSA generally focuses on the % survival at high DHA concentrations (700 nM) as there is minimal shift in the IC50 (see Figure 2), however, chloroquine clearly induces a shift in the IC50 (~5-fold), where the whole curve is shifted to the right, whereas the increase in % survival is relatively small. This different profile suggests that the mechanism of chloroquine-induced antagonism is different from the mechanism of artemisinin resistance. Current evidence regarding the mechanism of artemisinin resistance generally points towards decreased heme-mediated drug activation due to a decrease in hemoglobin uptake, which should be analogous to the decrease in heme-mediated drug activation caused by chloroquine. However, these different dose-response curves suggest different mechanisms are primarily responsible. Additional mechanisms have been proposed for artemisinin resistance, involving redox or heat stress responses, proteostatic responses, mitochondrial function, dormancy, and PI3K signaling, among others. Whilst the H-FluNox probe generally supports the idea that chloroquine suppresses heme-mediated DHA activation, it remains plausible that chloroquine could induce these, or other, cellular responses that suppress DHA activity.

The other potential weakness in the current manuscript is the interpretation of the OZ439 clinical data. Whilst the observed interaction with piperaquine and ferroquine may have been a contributing factor, it should also be recognised that the low pharmacokinetic exposure in these studies was the primary reason for treatment failure (Macintyre 2017).

Impact:

This study has important implications for the selection of drugs to form combinations for the treatment of malaria. The overall findings of antagonism between peroxide antimalarials and 4-aminoquinolines in the trophozoite stage are robust, and this carries across to the ring stage for chloroquine (but not piperaquine).

The manuscript also provides a plausible mechanism to explain the antagonism, although future work will be required to further explore the details of this mechanism and to rule out alternative factors that may contribute.

Overall, this is an important contribution to the field and provides a clear justification for the evaluation of potential drug combinations in relevant in vitro assays before clinical testing.

Reviewer #2 (Public review):

Summary:

This manuscript by Rosenthal and Goldberg investigates interactions between artemisinins and their quinoline partner drugs currently used for treating uncomplicated Plasmodium falciparum malaria. The authors show that chloroquine (CQ), piperaquine, and amodiaquine antagonize dihydroartemisinin (DHA) activity, and in CQ-resistant parasites, the interaction is described as "superantagonism," linked to the pfcrt genotype. Mechanistically, application of the heme-reactive probe H-FluNox indicates that quinolines render cytosolic heme chemically inert, thereby reducing peroxide activation. The work is further extended to triple ACTs and ozonide-quinoline combinations, with implications for artemisinin-based combination therapy (ACT) design, including triple ACTs.

Strengths:

The manuscript is clearly written, methodologically careful, and addresses a clinically relevant question. The pulsing assay format more accurately models in vivo artemisinin exposure than conventional 72-hour assays, and the use of H-FluNox and Ac-H-FluNox probes provides mechanistic depth by distinguishing chemically active versus inert heme. These elements represent important refinements beyond prior studies, adding nuance to our understanding of artemisinin-quinoline interactions.

Weaknesses:

Several points warrant consideration. The novelty of the work is somewhat incremental, as antagonism between artemisinins and quinolines is well established. Multiple prior studies using standard fixed-ratio isobologram assays have shown that DHA exhibits indifferent or antagonistic interactions with chloroquine, piperaquine, and amodiaquine (e.g., Davis et al., 2006; Fivelman et al., 2007; Muangnoicharoen et al., 2009), with recent work highlighting the role of parasite genetic background, including pfcrt and pfmdr1, in modulating these interactions (Eastman et al., 2016). High-throughput drug screens likewise identify quinoline-artemisinin combinations as mostly antagonistic. The present manuscript adds refinement by applying pulsed-exposure assays and heme probes rather than establishing antagonism de novo.

The dataset focuses on several parasite lines assayed in vitro, so claims about broad clinical implications should be tempered, and the discussion could more clearly address how in vitro antagonism may or may not translate to clinical outcomes. The conclusion that artemisinins are predominantly activated in the cytoplasm is intriguing but relies heavily on Ac-H-FluNox data, which may have limitations in accessing the digestive vacuole and should be acknowledged explicitly. The term "superantagonism" is striking but may appear rhetorical; clarifying its reproducibility across replicates and providing a mechanistic definition would strengthen the framing. Finally, some discussion points, such as questioning the clinical utility of DHA-PPQ, should be moderated to better align conclusions with the presented data while acknowledging the complexity of in vivo pharmacology and clinical outcomes.

Despite these mild reservations, the data are interesting and of high quality and provide important new information for the field.

Reviewer #3 (Public review):

Summary:

The authors present an in vitro evaluation of drug-drug interactions between artemisinins and quinoline antimalarials, as an important aspect for screening the current artemisinin-based combination therapies for Plasmodium falciparum. Using a revised pulsing assay, they report antagonism between dihydroartemisinin (DHA) and several quinolines, including chloroquine, piperaquine (PPQ), and amodiaquine. This antagonism is increased in CQ-resistant strains in isobologram analyses. Moreover, CQ co-treatment was found to induce artemisinin resistance even in parasites lacking K13 mutations during the ring-stage survival assay. This implies that drug-drug interactions, not just genetic mutations, can influence resistance phenotypes. By using a chemical probe for reactive heme, the authors demonstrate that quinolines inhibit artemisinin activation by rendering cytosolic heme chemically inert, thereby impairing the cytotoxic effects of DHA. The study also observed negative interactions in triple-drug regimens (e.g., DHA-PPQ-Mefloquine) and in combinations involving OZ439, a next-generation peroxide antimalarial. Taken together, these findings raise significant concerns regarding the compatibility of artemisinin and quinoline combinations, which may promote resistance or reduce efficacy.

Throughout the manuscript, no combinations were synergistic, which necessitates comparing the claims to a synergistic combination as a control. The lack of this positive control makes it difficult to contextualize the observed antagonism. Including a known synergistic pair (e.g., artemisinin + lumefantrine) throughout the study would have provided a useful benchmark to assess the relative impact of the drug interactions described.

Strengths:

This study demonstrates the following strengths:

(1) The use of a pulsed in vitro assay that is more physiologically relevant than the traditional 48h or 72h assays.

(2) Small molecule probes, H-FluNox, and Ac-H-FluNox to detect reactive cytosolic heme, demonstrating that quinolines render heme inert and thereby block DHA activation.

(3) Evaluates not only traditional combinations but also triple-drug combinations and next-generation artemisinins like OZ439. This broad scope increases the study's relevance to current treatment strategies and future drug development.

(4) By using the K13 wild-type parasites, the study suggests that resistance phenotypes can emerge from drug-drug interactions alone, without requiring genetic resistance markers.

Weaknesses:

(1) No combinations are shown as synergistic: it could be valuable to have a combination that shows synergy as a positive control (e.g, artemisinin + lumefantrine) throughout the manuscript. The absence of a synergistic control combination in the experimental design makes it more challenging to evaluate the relative impact of the described drug interactions.

(2) Evaluation of the choice of drug-drug interactions: How generalizable are the findings across a broad range of combinations, especially those with varied modes of action?

(3) The study would also benefit from a characterization of the molecular basis for the observed heme inactivation by quinolines to support this hypothesis - while the probe experiments are valuable, they do not fully elucidate how quinolines specifically alter heme chemistry at the molecular level.

(4) Suggestion of alternative combinations that show synergy could have improved the significance of the work.

(5) All data are derived from in vitro experiments, without accompanying an in vivo validation. While the pulsing assay improves physiological relevance, it still cannot fully capture the complexity of drug pharmacokinetics, host-parasite interactions, or immune responses present in living organisms.

(6) The absence of pharmacokinetic/pharmacodynamic modeling leaves questions about how the observed antagonism would manifest under real-world dosing conditions.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

We appreciate the positive assessment. We recognize that since all of the work in this manuscript was done in vitro, there are reasonable concerns about the translatability of these data to clinical settings. These results should not directly inform malaria policy, but we hope that these data bring new considerations to the approach for choosing strategic antimalarial combinations. We have modified the manuscript to clarify this distinction.

Public Reviews

Reviewer #1 (Public Review):

We thank the reviewer for their thoughtful summary of this manuscript. It is important to note that DHA-PPQ did show antagonism in RSAs. In this modified RSA, 200 nM PPQ alone inhibited growth of PPQ-sensitive parasites approximately 20%. If DHA and PPQ were additive, then we would expect that addition of 200 nM PPQ would shift the DHA dose response curve to the left and result in a lower DHA IC50. Please refer to Figure 4a and b as examples of additive relationships in dose-response assays. We observed no significant shift in IC50 values between DHA alone and DHA + PPQ. This suggests antagonism, albeit not to the extent seen with CQ. We have modified the manuscript to emphasize this point. As the reviewer pointed out, it is fortunate that despite being antagonistic, clinically used artemisinin-4-aminoquinoline combinations are effective, provided that parasites are sensitive to the 4-aminoquinoline. It is possible that superantagonism is required to observe a noticeable effect on treatment efficacy (Sutherland et al. 2003 and Kofoed et al. 2003), but that classical antagonism may still have silent consequences. For example, if PPQ blocks some DHA activation, this might result in DHA-PPQ acting more like a pseudo-monotherapy. However, as the reviewer pointed out, while our data suggest that DHA-PPQ and AS-ADQ are “non-optimal” combinations, the clinical consequences of these interactions are unclear. We have modified the manuscript to emphasize the later point.

While the Ac-H-FluNox and ubiquitin data point to a likely mechanism for DHA-quinoline antagonism, we agree that there are other possible mechanisms to explain this interaction. We have addressed this limitation in the discussion section. Though we tried to measure DHA activation in parasites directly, these attempts were unsuccessful. We acknowledge that the chemistry of DHA and Ac-H-FluNox activation is not identical and that caution should be taken when interpreting these data. Nevertheless, we believe that Ac-H-FluNox is the best currently available tool to measure “active heme” in live parasites and is the best available proxy to assess DHA activation in live parasites. These points are now addressed in the discussion section. Both in vitro and in parasite studies point to a roll for CQ in modulating heme, though an exact mechanism will require further examination. Similar to the reviewer, we were perplexed by the differences observed between in vitro and in parasite assays with PPQ and MFQ. We proposed possible hypotheses to explain these discrepancies in the discussion section. Interestingly, our data corelate well with hemozoin inhibition assays in which all three antimalarials inhibit hemozoin formation in solution, but only CQ and PPQ inhibit hemozoin formation in parasites. In both assays, in-parasite experiments are likely to be more informative for mechanistic assessment.

It remains unclear why K13 genotype influences RSA values, but not early ring DHA IC50 values. In K13WT parasites, both RSA values and DHA IC50 values were increased 3-5 fold upon addition of CQ. This suggests that CQ-mediated resistance is more robust than that conferred by K13 genotype. However, this does not necessarily suggest a different resistance mechanism. We acknowledge that in addition to modulating heme, it is possible that CQ may enhance DHA survival by promoting parasite stress responses. Future studies will be needed to test this alternative hypothesis. This limitation has been acknowledged in the manuscript. We have also addressed the reviewer’s point that other factors, including poor pharmacokinetic exposure, contributed to OZ439-PPQ treatment failure.

Reviewer #2 (Public Review):

We appreciate the positive feedback. We agree that there have been previous studies, many of which we cited, assessing interactions of these antimalarials. We also acknowledge that previous work, including our own, has shown that parasite genetics can alter drug-drug interactions. We have included the author’s recommended citations to the list of references that we cited. Importantly, our work was unique not only for utilizing a pulsing format, but also for revealing a superantagonistic phenotype, assessing interactions in an RSA format, and investigating a mechanism to explain these interactions. We agree with the reviewer that implications from this in vitro work should be cautious, but hope that this work contributes another dimension to critical thinking about drug-drug interactions for future combination therapies. We have modified the manuscript to temper any unintended recommendations or implications.

The reviewer notes that we conclude “artemisinins are predominantly activated in the cytoplasm”. We recognize that the site of artemisinin activation is contentious. We were very clear to state that our data combined with others suggest that artemisinins can be activated in the parasite cytoplasm. We did not state that this is the primary site of activation. We were clear to point out that technical limitations may prevent Ac-H-FluNox signal in the digestive vacuole, but determined that low pH alone could not explain the absence of a digestive vacuole signal.

With regard to the “reproducibility” and “mechanistic definition” of superantagonism, we observed what we defined as a one-sided superantagonistic relationship for three different parasites (Dd2, Dd2 PfCRTDd2, and Dd2 K13R539T) for a total of nine independent replicates. In the text, we define that these isoboles are unique in that they had mean ΣFIC50 values > 2.4 and peak ΣFIC50 values >4 with points extending upward instead of curving back to the axis. As further evidence of the reproducibility of this relationship, we show that CQ has a significant rescuing effect on parasite survival to DHA as assessed by RSAs and IC50 values in early rings.

Reviewer #3 (Public Review):

We thank the reviewer for their positive feedback. We acknowledge that no combinations tested in this manuscript were synergistic. However, two combinations, DHA-MFQ and DHA-LM, were additive, which provides context for contextualizing antagonistic relationships. We have previously reported synergistic and additive isobolograms for peroxide-proteasome inhibitor combinations using this same pulsing format (Rosenthal and Ng 2021). These published results are now cited in the manuscript.

We believe that these findings are specific to 4-aminoquinoline-peroxide combinations, and that these findings cannot be generalized to antimalarials with different mechanisms of action. Note that the aryl amino alcohols, MFQ and LM, were additive with DHA. Since the mechanism of action of MFQ and LM are poorly understood, it is difficult to speculate on a mechanism underlying these interactions.

We agree with the reviewer that while the heme probe may provide some mechanistic insight to explain DHA-quinoline interactions, there is much more to learn about CQ-heme chemistry, particularly within parasites.

The focus of this manuscript was to add a new dimension to considerations about pairings for combination therapies. It is outside the scope of this manuscript to suggest alternative combinations. However, we agree that synergistic combinations would likely be more strategic clinically.

An in vitro setup allows us to eliminate many confounding variables in order to directly assess the impact of partner drugs on DHA activity. However, we agree that in vivo conditions are incredibly more complex, and explicitly state this.

We agree that in the future, modeling studies could provide insight into how antagonism may contribute to real-world efficacy. This is outside the scope of our studies.

Recommendations for the Authors:

Reviewer #1 (Recommendations for the Authors):

The key weaknesses identified in this manuscript are described in the 'weaknesses' section of the public review. The major one is the inconsistency around the H-FluNox response in the chemical vs biological experiments. I can't think of a simple experiment to resolve this issue, but it is good that this data is openly provided in the manuscript. I believe there could be more discussion to clarify this limitation with the current study, and the conclusions, and particularly the title, should be softened regarding the mechanism of antagonism being based on heme reactivity.

We have softened the title and conclusions to take into account the limitations of our studies.

(1) Please double-check the definitions for isobologram interpretation. In most antimicrobial interaction studies, I see the threshold for antagonism at sumFIC50 of 1.5, or even 2. 1.25 is often interpreted as additive in many studies.

We acknowledge that different studies use various cutoff values. Our interpretations for additive versus antagonistic versus superantagonistic were based not only on mean ΣFIC50 values, but also isobologram shape. For example, the flat isoboles for MFQ-DHA were clearly distinct from the curved isoboles of PPQ-DHA. It is unclear what cutoff value(s) would be most clinically relevant.

(2) For the MFQ-PPQ interaction study, please make it clear that these drugs have very long half-lives (weeks), so the 4 h pulse assay isn't really relevant to their overall activity. It probably shows a slower onset of action, but there is plenty of drug remaining for many days in the clinical scenario, so perhaps the data from the traditional 48h assay is more relevant. The same consideration applies to OZ439, which may impact the interpretation of that data.

We have now included the half-lives of these compounds in the discussion section. Our intent was to use a pulsing format to make these isobolograms comparable with the other assays. It is important to note that pulses can reveal stronger phenotypes that might be missed with traditional methods. Thus, while 48 h assays may better mimic in vivo conditions, they could also mask important phenotypes.

Reviewer #3 (Recommendations for the Authors):

I have included most of my concerns in the public review. Below are some additional specific points for consideration:

(1) It is expected to include a synergistic combination as a control (e.g., artemisinin + lumefantrine) to contextualize the degree of antagonism observed. The experimental design should show some synergistic profiles in comparison. Adding a few experiments by including a synergistic control is needed.

Both MFQ-DHA and LM-DHA combinations were additive, which provides context for antagonistic combinations. This is now stated in the results section pertaining to Figure 1. We have also included a reference to our previous publication in which we demonstrated that proteasome inhibitor-peroxide combinations are synergistic to additive using this same pulsing format.

(2) Consider in vivo validation or pharmacokinetic/pharmacodynamic modeling to strengthen the translational relevance of the findings when it comes to doses and the IC50 correlations.

We agree that this would be useful to do in future, but it is outside the scope of the current study.

(3) It would be beneficial to include a discussion section on how the findings are generalizable to different Plasmodium falciparum genotypes (3D7, Dd2, MRA-1284) and their relevance.

Findings were consistent across three parasite backgrounds depending on PfCRT genotype. This point has been included in the discussion section. The background of these parasites is also provided in Table 1.

(4) Potential evaluation criteria to understand where certain combinations should be reconsidered can be included as a suggestion for the wider audience.

Our in vitro studies suggest that pulsing isobolograms would be a useful assay to include when evaluating combination therapies. While we believe that synergistic combinations would be more strategic than antagonistic combinations, we cannot provide evaluation criteria or make recommendations for reconsidering currently used combinations.

(5) Further elaborate on the mechanistic basis of heme inactivation by quinolines. If data are available, please include more data on the specificity of the process.

Despite our best efforts, we were unable to evaluate quinoline-heme interactions in parasites. Even in vitro, this interaction has remined elusive for decades. We agree that this would be an important future step towards supporting a specific mechanism for quinoline-DHA antagonism.

Version 2

Abstract

Artemisinin-based combination therapies (ACTs) remain the mainstay of treatment for Plasmodium falciparum malaria, despite reports of ACT treatment failure. ACTs consist of an artemisinin and a longer-lived partner drug, which is often a quinoline. Given that heme is central to the mechanism of action of artemisinins and some quinolines, we hypothesized that these antimalarials would exhibit strong drug-drug interactions. Previous studies using standard 48 h or 72 h assays identified additive to mildly antagonistic interactions between artemisinins and quinolines. Here, we sought to re-evaluate these interactions using a pulsing assay that better mimics the short in vivo half-life of artemisinins. We found that chloroquine (CQ), piperaquine (PPQ), and amodiaquine substantially antagonize dihydroartemisinin (DHA), the active metabolite of artemisinins. CQ-DHA antagonism was notably exacerbated in CQ-resistant parasites, resulting in a superantagonistic phenotype in isobolograms. Further, we found that CQ co-treatment conferred artemisinin resistance to Kelch 13 wild type parasites in the ring stage survival assay. Using a small molecule probe to measure chemically reactive heme in live parasites, we determined that quinolines block artemisinin activation by rendering cytosolic heme inert. Finally, we probed beyond traditional ACTs, evaluating interactions of the proposed triple ACT, DHA-PPQ-Mefloquine, as well as OZ439-quinoline combinations, which were all found to be antagonistic. Collectively, these in vitro data suggest that peroxide-quinolines may have liabilities as combination therapies.

Introduction

Malaria was responsible for an estimated 263 million cases and 597,000 deaths in 2023, the majority of which were caused by Plasmodium falciparum 1. The World Health Organization recommends artemisinin-based combination therapies (ACTs) for the treatment of P. falciparum malaria. ACTs are composed of a short-lived artemisinin derivative paired with a longer-lived partner drug. Depending on geographical resistance profiles, one of six combinations is recommended: Dihydroartemisinin (DHA) + Piperaquine (PPQ), Artesunate (AS) + Mefloquine (MFQ), AS + Amodiaquine (ADQ), AS + Pyronaridine (PYR), Artemether (AM) + Lumefantrine (LF), or AS + Sulfadoxine-pyrimethamine (SP). Three of these six partner drugs, PPQ, MFQ, and ADQ, possess quinoline groups. Importantly, heme is central to the mechanism of action of both artemisinins and quinolines 2,3.

Asexual blood stage parasites are adept at thriving in an incredibly heme rich environment. Parasites import up to 80% of host cell hemoglobin 4, which is transported to a specialized lysosome-like organelle, termed the digestive vacuole. Here, hemoglobin is proteolytically processed into globin peptides and redox active heme (Fe2+-ferroprotoporphyrin IX), which is oxidized to hemin (Fe3+-ferroprotoporphyrin IX) 5. The majority of hemin is retained within the digestive vacuole where it is detoxified via sequestration into inert hemozoin crystals 6–8. Hemin concentrations approach 0.5 M within the digestive vacuole 9. Blood stage parasites also maintain a relatively high cytosolic heme concentration of approximately 1.6 μM 10. While the source of this labile heme pool has not been experimentally determined, it is thought to originate from hemoglobin digestion.

The heme-rich environment of asexual blood stages has long been exploited for antimalarial intervention. The 4-aminoquinolines chloroquine (CQ) and PPQ are thought to act primarily in the parasite digestive vacuole, where they inhibit hemozoin formation by binding heme/hemin and the growing face of hemozoin 11–16. It has been proposed that disruption of heme homeostasis and/or accumulation of toxic heme-quinoline complexes may contribute to parasite killing 17–19. Resistance to 4-aminoquinolines is mediated by a collection of mutations in the chloroquine resistance transporter (PfCRT), which is located on the parasite digestive vacuole membrane 20. Mutations in PfCRT are thought to mediate resistance by enabling drug efflux out of the digestive vacuole 2,21–24.

Heme is also central to the mechanism of action of artemisinins. Artemisinins contain a peroxide bridge that must be cleaved by heme for antimalarial activity 25–29. Following activation, artemisinins non-specifically alkylate adjacent molecules, causing widespread cellular damage to the parasite 30–32. Clinically, artemisinin resistance is primarily associated with mutations in Kelch 13 (K13) 33–35, which are thought to reduce parasite hemoglobin uptake and digestion 36–38. Consequently, less heme is available to activate artemisinins.

ACTs remain the mainstay treatment for falciparum malaria. Widespread resistance to artemisinins in Southeast Asia and recent emergence of artemisinin resistance in Africa have posed great concern 1,34,39–42. Reduced artemisinin efficacy means that the partner drugs are responsible for clearing a larger parasite biomass, increasing the likelihood for resistance to emerge to partner drugs. Indeed, high treatment failure rates have been reported for AS-MFQ, DHA-PPQ, and AS-ADQ in Southeast Asia 43–51.

Given that heme is central to the mechanism of action of both quinolines and artemisinins, we hypothesized that these antimalarials would exhibit strong drug-drug interactions. Previously, several groups have assessed quinoline-artemisinin interactions using standard 48 or 72 h assays and found that combinations were additive in CQ-sensitive parasites and additive to slightly antagonistic in CQ-resistant parasites 52–56. However, standard 48 h or 72 h assays in which parasites are exposed to compounds for the entire assay duration do not well represent clinical drug exposure conditions and can mask not only resistance phenotypes, but also drug-drug interaction phenotypes 57–60. We sought to re-evaluate quinoline-artemisinin interactions in multidrug-sensitive, CQ-resistant, artemisinin-resistant, and PPQ-resistant parasites using pulsing assays that better represent artemisinin kinetics. Our work identifies that CQ, PPQ, and ADQ antagonize dihydroartemisinin (DHA), the active metabolite of clinically used artemisinins, likely by rendering heme chemically inert and preventing DHA activation. In addition, we explore drug-drug interactions beyond traditional ACTs, investigating interactions of putative triple ACTs as well as ozonide-quinoline combinations that failed in clinical trials. Overall, the data presented here highlight the importance of detailed investigation of drug-drug interactions, particularly in drug-resistant parasites.

Results

CQ and DHA are superantagonistic in CQ-resistant parasites

We began by evaluating CQ-DHA interactions in 3D7, a CQ-sensitive parasite, and Dd2, a CQ-resistant parasite (Table 1), using trophozoite-stage isobologram assays. Synchronized trophozoites were treated with fixed ratios of CQ and DHA for 4 h, drugs were washed off, and then parasitemia was assessed 72 h later. Given the short half-life of artemisinins 58, we believe that this pulsing format allows us to better assess the effect of ACT partner drugs on DHA activity compared to standard 48 or 72 h assays. Fractional IC50 (FIC50) values were calculated from fixed ratios and then plotted on an XY-graph to generate isobolograms. Isobolograms can be interpreted based on isobole shape and mean ΣFIC50 values as follows: points along the dotted line with a mean ΣFIC50 value close to 1 indicate additivity, points above the dotted line in a concave curve with a mean ΣFIC50 value >1.25 indicate antagonism, and points below the dotted line in a convex curve with a mean ΣFIC50 value <0.75 indicate synergy.

We observed strong antagonism for CQ-DHA in 3D7 parasites (Figure 1a), with a mean ΣFIC50 value of 1.68. Antagonism was notably exacerbated in Dd2 parasites (Figure 1b), which had a mean ΣFIC50 value of 2.44. The magnitude of this value far surpasses conventional antagonism, and we termed this phenomenon “superantagonism”. Perhaps even more unexpected was the atypical shape of this isobole (Figure 1b), where points extended upward and peak ΣFIC50 values exceeded 4. In contrast, peak ΣFIC50 values for 3D7 reached only 2.

CQ and DHA are superantagonistic in CQ-resistant parasites.

Shown are trophozoite-stage isobolograms for (a-e) CQ-DHA, (f-k) PPQ-DHA, and (l-q) MFQ-DHA. The dotted line on each graph represents perfect additivity. 3D7, Dd2, Dd2 PfCRT3D7, Dd2 PfCRTDd2, Dd2 K13R539T, and Dd2 PfCRTM343L parasites are indicated in orange, grey, pink, blue, purple, and teal, respectively. Data from three independent replicates are indicated by different shading. Figure supplement 1. CQ, DHA, PPQ, and MFQ sensitivity of different parasites.

The traditional isobole shape for 3D7 indicates a reciprocal relationship between CQ and DHA, whereas the atypical isobole shape of Dd2 suggests a one-sided relationship, whereby CQ is blocking DHA activity.

Multiple groups have demonstrated that parasite genetics can influence drug-drug interactions 55,61,62. 3D7 and Dd2 parasites differ genetically at a number of drug resistance markers, including PfCRT which is responsible for mediating CQ resistance. To determine if PfCRT genotype was responsible for differences in CQ-DHA interactions between these parasite lines, we evaluated interactions in two Dd2 parasites that are isogenic except at PfCRT 63,64: Dd2 PfCRT3D7, which is edited to have the 3D7 PfCRT genotype, and Dd2 PfCRTDd2, which maintains the Dd2 PfCRT genotype and was used as a cloning control (Table 1). Notably, superantagonism was lost in Dd2 PfCRT3D7 parasites, which phenocopied 3D7 parasites, and had a mean ΣFIC50 value of 1.85 and a conventional isobole shape (Figure 1c). In contrast, Dd2 PfCRTDd2 parasites displayed one-sided superantagonism with a mean ΣFIC50 value of 2.73 and peak ΣFIC50 values > 4.5 (Figure 1d). These data suggest that PfCRT genotype is the main genetic determinant that dictates CQ-DHA interactions.

Both 3D7 and Dd2 parasites are artemisinin sensitive, harboring a wildtype K13. To determine if resistance to artemisinins alters CQ-DHA interactions, we next evaluated CQ-DHA combinations in Dd2 parasites that are edited to have a K13R539T mutation 35 (Table 1). Dd2 K13R539T parasites, which also harbor a Dd2 PfCRT genotype, displayed a one-sided superantagonistic isobole that was reminiscent of both Dd2 and Dd2 PfCRTDd2 parasites (Figure 1e). This suggests that, unlike PfCRT genotype, K13 genotype appeared to have no impact on drug-drug interactions at the trophozoite stage.

PPQ and DHA are antagonistic

Next, we sought to evaluate drug-drug interactions of DHA and the ACT partner drugs PPQ and MFQ. PPQ is believed to have a similar mechanism of action as CQ and both antimalarials are 4-aminoquinolines. Accordingly, we hypothesized that PPQ and DHA would also be antagonistic. In 3D7, Dd2, Dd2 PfCRT3D7, Dd2 PfCRTDd2, and Dd2 K13R539T trophozoites, we observed conventional antagonism for PPQ-DHA (Figure 1f-j) with mean ΣFIC50 values ranging from 1.55 to 1.71. To determine if, similar to CQ, PPQ-DHA interactions would be exacerbated in PPQ-resistant parasites, we also evaluated Dd2 PfCRTM343L, which harbors the Dd2 PfCRT plus an additional mutation at M343L 64 (Table 1). The M343L mutation had no impact on PPQ-DHA interactions and Dd2 PfCRTM343L parasites displayed conventional antagonism with a mean ΣFIC50 value of 1.58 (Figure 1k). Note that while the PPQ IC50 of Dd2 PfCRTM343L is 2.5-fold higher than Dd2 (65 nM versus 25 nM; Figure 1-figure supplement 1), this shift is substantially smaller than the >1000-fold CQ IC50 shift that we and others 60 have observed in CQ-sensitive versus CQ-resistant parasites using the pulsing assay format (Figure 1-figure supplement 1).

Next, we assessed interactions between MFQ and DHA (Figure 1l-q). Though MFQ contains a quinoline group, it is chemically classified as an aryl amino alcohol and is believed to have a heme-independent mechanism of action 65. Consistent with this, we observed additivity for MFQ-DHA in all six parasites tested with mean ΣFIC50 values ranging from 0.94 to 1.20. This additivity along with the synergism that we previously reported for DHA-proteasome inhibitor combinations 61, provides context for the antagonism and superantagonism that we observed for PPQ-DHA and CQ-DHA.

CQ combination treatment confers artemisinin resistance to K13WT parasites in ring-stage survival assays

Given our observation that CQ and PPQ antagonize DHA in trophozoite stages, we next sought to determine if these quinolines would promote survival in early ring stages where artemisinin resistance is classically observed. We performed DHA dose-response assays on early ring stage parasites to determine both ring-stage survival assay (RSA) values and IC50 values (Figure 2a and Figure 2-figure supplement 1). The RSA is used to delineate artemisinin-sensitive versus resistant parasites, where survival > 1% corresponds to resistance and < 1% indicates sensitivity 58. Though DHA IC50 values cannot reliably differentiate artemisinin sensitive vs resistant parasites 58, they can provide important insight into different factors that contribute to artemisinin susceptibility 66.

CQ co-treatment confers artemisinin resistance in K13WT parasites.

(a) DHA dose response assays were performed on 0-3 hours post invasion (hpi) ring stage parasites. Two-fold serial dilutions of DHA were prepared in 96-well plates (DHA concentration represented by different shades of purple) starting at 1.4 μM so that both IC50 values and RSA survival values at 700 nM DHA could be determined. To evaluate quinoline-DHA interactions in early ring stages, DHA titrations were prepared with fixed concentrations of 10 μM CQ or 200 nM PPQ. (b-d) Mean RSA survival values ± SEM (top panel) and mean DHA IC50 values ± SEM (bottom panel) are shown for (b) Dd2 PfCRT3D7, (c) Dd2, and (d) Dd2 K13R539T parasites. For the PPQ-resistant parasites (e) Dd2 PfCRTM343L and (f) MRA-1284, additional fixed PPQ concentrations of 500 nM and 10 μM were also tested. For b bottom panel, a student’s t-test was used to assess statistical differences between DHA alone versus DHA + PPQ. For all other graphs, statistical differences between DHA alone (-) and DHA + quinolines was assessed using a one-way ANOVA with a Dunnett’s test for multiple comparisons. ****p < 0.0001; ***p < 0.001; *p < 0.05; ns = not significant. Figure supplement 1. Early ring stage dose-response curves

In the absence of quinolines, the artemisinin sensitive K13WT parasites, Dd2 PfCRT3D7 and Dd2, had RSA survival values ≤ 1%, while Dd2 K13R539T had an RSA survival of 12.63% (Figure 2b-d top panel). Co-treatment with 10 μM CQ, a pharmacologically relevant concentration 67,68, significantly increased RSA survival, conferring artemisinin resistance to both Dd2 PfCRT3D7 and Dd2 parasites with mean RSA values of 2.66% and 4.49%, respectively (Figure 2b and c top panel). RSA survival values were also increased in Dd2 K13R539T parasites to 18.39%, although this shift was not statistically significant (Figure 2d top panel). When assessing DHA sensitivity by ring-stage IC50 values, CQ co-treatment also had a substantial rescuing effect in both Dd2 and Dd2 K13R539T parasites, resulting in a 4-fold increase in DHA IC50 (Figure 2c and d bottom panel). Note that while we were unable to determine a DHA IC50 value for CQ-sensitive Dd2 PfCRT3D7 parasites in the presence of 10 μM CQ, the dose-response curve was suggestive of antagonism (Figure 2-figure supplement 1a).

To determine if PPQ would have a similar rescuing effect, we performed these same assays in the presence of 200 nM PPQ. This concentration was chosen because it is sublethal for early rings in this short 3 h pulse, pharmacologically relevant 69,70, and mimics the PPQ survival assay (PSA) that is used to delineate PPQ resistance 57. Co-treatment with 200 nM PPQ did not have a rescuing effect for Dd2 PfCRT3D7, Dd2, or Dd2 K13R539T parasites and resulted in mean RSA values that were either similar to or lower than the DHA-only control (Figure 2b-d). Note that in the absence of DHA, 200 nM PPQ inhibited parasite growth approximately 20%. However, co-treatment with 200 nM PPQ did not shift the IC50 values or DHA dose response curves of these parasites (Figure 2b-d and Figure 2-figure supplement 1a-c), indicating that PPQ and DHA are also antagonistic in early ring stages.

We were curious if higher concentrations of PPQ might have a rescuing effect, particularly for PPQ-resistant parasites that are capable of surviving a higher PPQ dose. To this end, we assayed two PPQ-resistant parasites: Dd2 PfCRTM343L and MRA-1284 (IPC_6261), which is an artemisinin-resistant and PPQ-resistant clinical isolate that was collected in 2012 from Cambodia (Table 1). Dd2 PfCRTM343L and MRA-1284 have PPQ survival values (PSA) of 10% 64 and 40% 71, respectively. While co-treatment with 10 μM CQ had a rescuing effect for Dd2 PfCRTM343L as assessed by RSA and IC50 values, neither 200 nM PPQ, 500 nM PPQ, or 10 μM PPQ co-treatment rescued parasite survival (Figure 2e and Figure 2-figure supplement 1d). Similarly, 10 μM CQ co-treatment decreased MRA-1284 sensitivity to DHA as assessed by RSA and early ring IC50 values, while co-treatment with 200 nM PPQ or 500 nM PPQ had no effect on DHA sensitivity (Figure 2f). This phenotype is consistent with conventional antagonism. Interestingly, while co-treatment with 10 μM PPQ did not alter the RSA survival of MRA-1284, it did shift the dose-response curve to the left (Figure 2-figure supplement 1e). However, since 10 μM PPQ inhibited parasite growth approximately 60% on its own, we were unable to determine an IC50 value.

CQ and PPQ interfere with heme iron reactivity

In total, we profiled three different quinoline-DHA interactions, which are summarized in Figure 3a. In solution, quinolines and heme form heme-quinoline complexes, wherein heme is present as a µ-oxo dimer 72–74. Though the exact stoichiometry of this interaction for each quinoline remains debated 72–76, formation of such a complex could render quinoline-bound heme chemically inert (i.e. unable to activate artemisinins).

CQ and PPQ antagonize DHA activation by “inactivating” heme.

(a) Shown are mean ΣFIC50 values calculated from the trophozoite stage isobolograms in Figure 1. Values close to 1 (pink) indicate additivity. Values ranging from 1.25 to 2 (light blue) indicate classical antagonism. Values > 2.25 (dark turquoise) indicate superantagonism. (b top panel) Heme (Fe2+-FPIX) cleaves the N-O bond of H-FluNox, mediating probe fluorescence. (b bottom panel) Heme cleaves the peroxide bridge of DHA, which generates a carbon centered radical that can alkylate proximal biomolecules. (c) H-FluNox was incubated with increasing concentrations of heme in the presence or absence of 10 μM CQ, 150 nM PPQ, 10 μM PPQ, 150 nM MFQ, or 10 μM MFQ. Shown are mean relative fluorescence units (RFU) ± SD from at least three independent replicates performed in technical triplicate. (d and e) Dd2 trophozoites were treated with 10 μM CQ, 150 nM PPQ, 150 nM MFQ, or mock treated for 5.5 h. (f) Dd2 PfCRTDd2 and Dd2 PfCRT3D7 trophozoites were mock treated or treated with 150 nM CQ or 10 μM CQ for 5.5 h. (d-f) Parasites were then incubated with 10 μM Ac-H-FluNox (the cell permeable analog of H-FluNox) to visualize and quantify “active” heme. Shown are (d) representative images and (e and f) mean fluorescence intensity ± SD of at least 70 parasites from at least three independent drug treatments. Statistical significance was assessed using a one-way ANOVA with a Turkey’s test for multiple comparisons. (g and h) Dd2 trophozoites were treated with 10 μM CQ, 25 nM DHA, 10 μM CQ + 25 nM DHA for 6 h. Parasite lysates were subjected to western blot and probed with anti-K48-linked ubiquitin (K48-Ub) antibodies or anti-plasmepsin V (PMV) antibodies. Shown is a (g) representative blot and (h) quantification of relative K48-Ub intensity ± SEM from three independent replicates. Statistical significance between mock treated and antimalarial-treated parasites was determined using a one-way ANOVA with a Turkey’s test for multiple comparisons. ****p < 0.0001; ***p < 0.001; **p < 0.01; ns = not significant. Figure supplement 1. Effect of pH on H-FluNox and Ac-H-FluNox activity. Figure supplement 2. Equimolar comparison of CQ, PPQ, and MFQ on Ac-H-FluNox fluorescence. Figure supplement 3. Representative images of Dd2 PfCRTDd2 and Dd2 PfCRT3D7 Ac-H-FluNox fluorescence. Figure supplement 4. Additional K48-ubiquitin western blots. Table supplement 1. Mean parasite intensity for Ac-H-FluNox experiments. Source data 1. Uncropped western blots used for ImageJ quantification.

To investigate this, we utilized a recently developed small molecule probe, H-FluNox, that is highly sensitive and specific for heme over hemin, heme-bound proteins, Fe2+, and other salts and metal ions 77. H-FluNox is basally non-fluorescent, but upon cleavage of a N-O bond by the Fe2+ of heme, H-FluNox fluoresces 77 (Figure 3b top panel). While not identical, this reaction is similar to artemisinin activation, which requires cleavage of artemisinin’s peroxide bond by the Fe2+ of heme 29 (Figure 3b bottom panel). To determine if quinoline-heme complexes would react with H-FluNox, we began by assessing H-FluNox fluorescence in solution with 10 μM CQ, which is sub-lethal for CQ-resistant parasites and pharmacologically relevant 67,68. Relative to the untreated control, 10 μM CQ inhibited fluorescence approximately 50% in the presence of 250 nM heme (Figure 3c). This suggests that heme iron reactivity is impaired when heme is in complex with CQ. Next, we evaluated PPQ and MFQ at two different concentrations: 150 nM, which reflects a 5x trophozoite-stage IC50 concentration, and 10 μM to directly compare these quinolines with CQ. At 150 nM, both PPQ and MFQ inhibited fluorescence, albeit only 10% (Figure 3c). At 10 μM, MFQ inhibited fluorescence to a similar extent as CQ, while 10 μM PPQ only inhibited fluorescence 25% (Figure 3c). Note that 10 μM concentrations exceed maximum plasma concentrations of PPQ and MFQ, which have been reported to reach approximately 0.5-1 μM and 5 μM, respectively 69,70,78.

Next, we assessed fluorescence in live trophozoite stage parasites using the cell permeable analog of H-FluNox, Ac-H-FluNox 77. In untreated Dd2 parasites, fluorescence was predominantly observed in the parasite cytoplasm (Figure 3d), and little signal was detected in red blood cells or the parasite digestive vacuole. This suggests that a significant portion of DHA could be activated in the parasite cytoplasm. Lack of signal in the digestive vacuole could be due to issues with permeability, the specificity of Ac-H-FluNox for heme over hemin (the predominant heme species in the digestive vacuole), or sensitivity of Ac-H-FluNox to a low pH environment. In solution, H-FluNox fluorescence was decreased approximately 60% at pH 5.4 (Figure 3-figure supplement 1a), which is the estimated pH of the digestive vacuole 79. However, alkalinization of the digestive vacuole with ammonium chloride 80 did not increase digestive vacuole fluorescence signal (Figure 3-figure supplement 1b), suggesting that low pH alone cannot explain this lack of labeling.

Importantly, treatment of Dd2 trophozoites with 10 μM CQ nearly ablated Ac-H-FluNox fluorescent signal (Figure 3d and e). Treatment with 150 nM PPQ also significantly reduced fluorescence, though not to the same extent as 10 μM CQ (50% inhibition versus > 92% inhibition) (Figure 3d, e, and Figure 3 and 4-table supplement 1). In contrast, parasites treated with 150 nM MFQ had similar fluorescence as untreated control parasites (Figure 3d and e). These assays with H-FluNox and Ac-H-FluNox correlate well with hemozoin inhibition assays in which CQ, PPQ, and MFQ inhibit hemozoin formation in solution 73, but only CQ and PPQ inhibit hemozoin formation in parasites 7,81–83. To determine if concentration alone was responsible for the differences that we observed between CQ and the other quinolines, we also evaluated Ac-H-FluNox fluorescence following treatment with 10 μM PPQ or 10 μM MFQ. At this super-physiological and super-lethal concentration, an equimolar concentration of CQ still inhibited H-FluNox fluorescence to a greater extent than PPQ or MFQ (Figure 3-figure supplement 2 and Figure 3 and 4-table supplement 1). This suggests that something unique to the chemistry of CQ and/or the biology of the parasite is responsible for the differences that we observed.

Quinolines antagonize peroxides.

(a and b) Trophozoite-stage DHA dose-response assays were performed with Dd2 parasites in the presence or absence of 15 nM PPQ, 15 nM MFQ, 15 nM PPQ + 15 nM MFQ, or 2 nM DHA. Shown are (a) dose-response curves with mean % inhibition ± SEM and (b) mean IC50 values ± SEM from at least three independent replicates. Statistical significance was determined using a one-way ANOVA with a Turkey’s test for multiple comparisons. Trophozoite-stage isobolograms were performed on Dd2 parasites to determine drug-drug interactions of (c) PPQ-MFQ, (d) ADQ-DHA, (e) LM-DHA, (f) PPQ-OZ439, and (g) FQ-OZ439 using the following fixed ratios: 1:0, 4:1, 2:1, 1:1, 1:2, 1:4, 0:1. Shown are fractional IC50 values from at least 3 independent replicates. Independent replicates are indicated in different shading and the diagonal dotted line on each plot indicates perfect additivity. (h and i) Dd2 parasites were mock treated or treated with 10 μM CQ, 150 nM ADQ, 500 nM LM, or 150 nM FQ for 5.5 h and free heme was then labeled in live parasites with Ac-H-FluNox. Shown are (h) representative images and (i) mean fluorescence intensity ± SD of at least 90 parasites from three independent drug treatments. Statistical significance was assessed using a one-way ANOVA with a Dunnett’s test for multiple comparisons. ****p < 0.0001; ns = not significant. (j) Mean relative fluorescence of Ac-H-FluNox from drug treatments in Figure 3e and 4i was plotted against Dd2 mean ΣFIC50 values for the indicated combinations. Combinations with DHA are indicated in black. Combinations with OZ439 are indicated in blue. Figure supplement 1. Additional MFQ isobolograms. Figure supplement 2. ADQ, LM, and FQ IC50 values. Table supplement 1. Mean parasite intensity for Ac-H-FluNox experiments.

PfCRTDd2 is believed to mediate CQ resistance by pumping CQ out of the digestive vacuole 2,21–24, suggesting that cytoplasmic CQ concentrations may differ between PfCRTDd2 and PfCRT3D7 parasites treated with the same concentration of CQ. Given the substantial difference that we saw in the shape of CQ-DHA isobolograms between PfCRTDd2 and PfCRT3D7 parasites, we were curious if differences in PfCRT genotype would impact Ac-H-FluNox fluorescence. Accordingly, we treated Dd2 PfCRTDd2 and Dd2 PfCRT3D7 parasites with 150 nM CQ or 10 μM CQ and measured chemically “reactive” heme using Ac-H-FluNox. In untreated Dd2 PfCRTDd2 versus Dd2 PfCRT3D7 parasites, we observed no difference in Ac-H-FluNox localization or fluorescence intensity (Figure 3f and Figure 3-figure supplement 3). Following treatment with 150 nM CQ, fluorescence was inhibited 25% in Dd2 PfCRT3D7 and 35% in Dd2 PfCRTDd2, although this difference was not statistically significant (Figure 3f and Figure 3 and 4-table supplement 1). Curiously, while 10 μM CQ treatment reduced mean fluorescence approximately 60% in Dd2 PfCRT3D7 parasites, it was not to the extent of Dd2 PfCRTDd2 parasites (80%) (Figure 3f and Figure 3 and 4-table supplement 1). These data suggest that transport of CQ out of the digestive vacuole by PfCRTDd2 may contribute to superantagonism between CQ and DHA in CQ-resistant parasites.

CQ co-treatment rescues DHA-mediated protein damage

Following activation by heme, DHA causes widespread protein damage in the parasite 30–32,84. Accordingly, we would expect that if CQ is antagonizing DHA activation, then CQ co-treatment should rescue DHA-mediated protein damage. To test this hypothesis, we treated Dd2 trophozoites with CQ, DHA, or CQ + DHA and probed for K48-linked ubiquitin, which is a marker of protein damage 84. Consistent with what has been previously reported 85, CQ treatment alone had no effect on K48-linked ubiquitin levels, while treatment with DHA resulted in a 2-fold increase in K48-linked ubiquitin (Figure 3g, h and Figure 3-figure supplement 4). Importantly, addition of CQ to DHA had a rescuing effect, resulting in levels of K48-linked ubiquitin that were similar to the DMSO-treated control (Figure 3g and h). Collectively, these data suggest that CQ antagonizes DHA by rendering heme chemically inert, reducing DHA activation, and consequently, reducing DHA-mediated protein damage.

PPQ and MFQ do not harm treatment efficacy, but strongly antagonize each other

With the looming threat of widespread ACT treatment failure, several ideas have been proposed to delay the spread of ACT resistance by using currently available antimalarials. Triple ACTs (TACTs), which are composed of a traditional ACT plus an additional longer-lived partner drug, have gained the most traction. Early clinical trials indicate that DHA-PPQ-MFQ and AM-LM-ADQ show good safety and efficacy against artemisinin-sensitive and resistant parasites 47,86–88. We were curious how the addition of a third antimalarial would influence drug interactions. Accordingly, we performed trophozoite-stage DHA dose response assays on Dd2 parasites in the presence of a fixed, sub-lethal concentration of PPQ, MFQ, or PPQ + MFQ. Addition of 2 nM DHA was also included as a control to represent additivity. As assessed by DHA IC50 values, addition of 15 nM PPQ did not improve parasite killing, while addition of 15 nM MFQ resulted in a 3-fold decrease in IC50 compared to the DMSO control (Figure 4a and b). These phenotypes are consistent with antagonistic and additive relationships, respectively. Notably, addition of PPQ + MFQ had no additional benefit compared to addition of MFQ alone (Figure 4 a and b).

Following the initial 3-day TACT treatment regimen, long-lived partner drugs are responsible for clearing the remaining parasite burden. Accordingly, we sought to assess drug-drug interactions of PPQ-MFQ in the absence of DHA. Previously it was reported that MFQ and PPQ were moderately antagonistic in 3D7 and K1 parasites using a standard 48 h assay, with mean ΣFIC50 values of 1.31 and 1.28, respectively 52. Using our trophozoite-stage assay, we observed reciprocal superantagonism in both PPQ-sensitive, Dd2 (Figure 4c), and PPQ-resistant Dd2 PfCRTM343L (Figure 4-figure supplement 1a), with mean ΣFIC50 values of 2.62 and 2.44, respectively. This phenotype appears to be unique to PPQ, as we observed only strong antagonism for CQ-MFQ with a mean ΣFIC50 of 1.72 (Figure 4-figure supplement 1b). Overall, these in vitro data suggest that based on drug-drug interactions, PPQ and MFQ may not be a strategic long-lived antimalarial pairing for TACTs.

Quinolines antagonize peroxides

To determine if the drug-drug interactions that we evaluated were specific to a subset of antimalarials or represented a broader phenotype, we next evaluated DHA interactions with the clinically used ACT partner drugs ADQ and LM. Structurally, ADQ is classified as a 4-aminoquinoline like PPQ and CQ, while LM is classified as an aryl amino alcohol like MFQ.

However, unlike MFQ, LM lacks a quinoline group. We found that in Dd2 parasites, ADQ and DHA were antagonistic (Figure 4d) with a mean ΣFIC50 of 1.61, while LM and DHA were additive (Figure 4e) with a mean ΣFIC50 of 1.02.

To determine if 4-aminoquinolines antagonize other peroxide antimalarials, we next sought to assess interactions of two different OZ439-quinoline pairs that were previously evaluated in clinical trials. Similar to artemisinins, the synthetic ozonide, OZ439, requires activation via heme-mediated cleavage of a peroxide bond, but has a significantly improved half-life of 46-62 h 89. In phase II clinical trials, OZ439 showed good efficacy alone 89, but poor efficacy either in combination with PPQ 90 or the next generation quinoline, ferroquine (FQ) 91,92. To assess drug-drug interactions, we performed trophozoite-stage isobolograms with OZ439-PPQ and OZ439-FQ in Dd2 parasites. The pulsing format was used to maintain consistency across experiments. Both PPQ-OZ439 (Figure 4f) and FQ-OZ439 (Figure 4g) were antagonistic with mean ΣFIC50 values of 1.42 and 1.52, respectively. These data suggest that antagonistic interactions are potentially one of several factors 90–94 that may have contributed to the poor efficacy of these single dose combination therapies.

Although the mechanism of action of ADQ and FQ are less well studied, both antimalarials appear to inhibit hemozoin formation. ADQ was shown to inhibit hemozoin formation in parasites using the heme fractionation assay 83. While in-parasite assays have not been performed with FQ, in vitro biochemical studies suggest that FQ may inhibit hemozoin formation even better than CQ 95. Accordingly, we were curious if ADQ and FQ treatment would “inactivate” heme in parasites. Using the Ac-H-FluNox heme probe, we observed that relative to the untreated control, treatment with 150 nM ADQ or 150 nM FQ (approximate 5x IC50 concentrations; Figure 4-figure supplement 2) reduced Ac-H-FluNox fluorescent signal approximately 50% and 55%, respectively (Figure 4h, i, and Figure 3 and 4-table supplement 1). In contrast, treatment with an equipotent concentration of LM had no significant impact on Ac-H-FluNox signal relative to the untreated control (Figure 4h and i). Among the six partner drugs that were evaluated, we observed a clear inverse correlation between Ac-H-FluNox signal and mean ΣFIC50, with an R2 value of 0.84 (Figure 4j). These data strongly suggest that antimalarials that render heme inert also antagonize artemisinin and ozonide antimalarials, presumably by blocking endoperoxide activation.

Discussion

Combination therapies are widely believed to delay the emergence of drug resistance not only for malaria, but also bacterial infections, HIV, fungal infections, and cancer. However, to maximize the benefits of combination therapies, it is essential that drugs are strategically paired. At a minimum, combinations should enhance efficacy to reduce parasite biomass and reduce the potential for drug resistance to evolve. In vitro assessment of drug-drug interactions between two antimalarials with vastly different half-lives is challenging. We chose to use a short treatment pulse to mimic artemisinin exposure and gain an understanding into how quinolines impact artemisinin activity. While drug-drug interactions are undoubtably much more complex in vivo, we believe this assay format better recapitulates clinical conditions compared to traditional 48 or 72 h assays.

We found that CQ and PPQ are antagonistic with DHA as assessed in early ring and trophozoite stages across three different parasite backgrounds. Artemisinin resistance is thought to be mediated by reduced heme-mediated activation and/or enhanced parasite stress response. Accordingly, 4-aminoquinolines could antagonize DHA by either or both mechanism(s). Our work with Ac-H-FluNox suggests that 4-aminoquinolines antagonize DHA by rendering intra-parasitic heme chemically inert, thus blocking DHA activation. However, future work will be needed to assess the effect of 4-aminoquinolines on parasite stress responses.

To date, it remains debated where artemisinins are activated within the parasite. There are several lines of evidences that suggest that artemisinins can be activated in the parasite cytoplasm. First, immunofluorescence studies with an anti-AS antibody 96 and live cell imaging with an artemisinin-based photoaffinity probe 97 revealed that artemisinins localize throughout the parasite cytoplasm. Second, numerous chemical proteomics studies utilizing artemisinin and ozonide-based click chemistry probes have identified a number of cytoplasmic proteins as targets of artemisinin and ozonides 30–32,98,99. Given how quickly activated artemisinins react with adjacent molecules, it is unlikely that activated artemisinins could traverse the digestive vacuole or other organelle membranes before encountering a cytoplasmic target. Finally, immunofluorescence assays with antibodies specific for the alkylation signature of the ozonide OZ277 100 revealed labeling throughout the parasite cytoplasm. Our data with Ac-H-FluNox also lend credence to this hypothesis. We observed Ac-H-FluNox fluorescence throughout the parasite cytoplasm, indicating that there is a substantial amount of heme in this location that is chemically competent to activate artemisinins. It is possible that technical limitations, such as issues with membrane permeability, limit Ac-H-FluNox fluorescence in the digestive vacuole. Nevertheless, these data collectively indicate that artemisinins can be activated in the parasite cytoplasm.

Previously, it has been reported that CQ treatment increases intracellular “free” heme 7,10,101. However, the term “free” heme is somewhat malleable. While the pyridine-heme fractionation assay 7,83, electron spectroscopic imaging (ESI) with energy-loss spectroscopy (EELS) 7 and heme biosensor 10, can be used to detect and quantify heme species, they are unable to differentiate chemically “active” heme from inert heme (i.e. quinoline bound or protein bound). In contrast, H-FluNox is highly specific for chemically “active” heme 77. One possibility to reconcile our data with previous reports is that CQ treatment increases non-hemozoin/hemoglobin heme, but that this heme is bound to CQ or other biomolecules, rendering it unable to activate Ac-H-FluNox or peroxides. It is important to note that the chemistry of artemisinin and Ac-H-FluNox activation are not identical and these data with Ac-H-FluNox should be interpreted with caution. Nevertheless, we believe that Ac-H-FluNox is the best currently available proxy to assess artemisinin activation in live parasites.

CQ-DHA antagonism was notably exacerbated in CQ-resistant parasites. The concentrations used in isobolograms are chosen based on parasite IC50 values to individual compounds. While all parasites were exposed to similar DHA concentrations, CQ-resistant parasites were exposed to much higher CQ concentrations compared to their CQ-sensitive counterparts. This may explain in part why antagonism was intensified in CQ resistant parasites. However, equimolar treatment with 10 µM CQ resulted in significantly lower Ac-H-FluNox signal in Dd2 PfCRTDd2 versus Dd2 PfCRT3D7 parasites (Figure 3f). These data could suggest that while the majority of CQ is sequestered in the digestive vacuole of Dd2 PfCRT3D7 parasites, PfCRTDd2 is able to transport more CQ out of the digestive vacuole, thus allowing more CQ to bind cytosolic heme in parasites harboring a PfCRTDd2. This duality of PfCRTDd2, simultaneously enhancing parasite survival to CQ while promoting cytoplasmic heme “inactivation”, may explain the unprecedented one-sided superantagonistic isobole that we observed.

Fortunately, we did not observe superantagonism with DHA-PPQ in PPQ-resistant parasites. With a modest 3-fold shift in PPQ IC50 between Dd2 vs Dd2 PfCRTM343L parasites (25 nM vs 65 nM), it is perhaps unsurprising that no significant difference in DHA-PPQ isobolograms was observed between these two parasites. This is in contrast to the >1000-fold CQ IC50 shift that we and others 60 have observed between CQ-sensitive and CQ-resistant parasites using pulsing assays. Our studies with Ac-H-FluNox indicate that 150 nM PPQ is sufficient to render approximately 50% of available heme inert (Figure 3e). While these data suggest that PPQ blocks some DHA activation, it is likely that 150 nM PPQ is not sufficient to block all DHA activation. Since even PPQ-resistant parasites are at least partially susceptible to mid-nanomolar PPQ concentrations, the protective effect that PPQ has in neutralizing DHA is partially negated by PPQ-mediated killing. In contrast, CQ-resistant parasites are relatively unaffected by pulses with low micromolar CQ concentrations. Accordingly, CQ concentrations that are high enough to nearly block all DHA activation are insufficient to kill CQ resistant parasites.

Relative to PPQ and MFQ, CQ appears to have the greatest ability to impair heme iron reactivity. We were surprised to observe that PPQ inhibited H-FluNox fluorescence only half as well as CQ in solution. This suggests that there may be important differences in how these 4-aminoquinolines bind heme. In Dd2 parasites, CQ also inhibited Ac-H-FluNox fluorescence to a greater extent than PPQ at 10 μM concentrations. Given that the ability of CQ to “inactivate” heme in parasites is dependent on PfCRT genotype, it is possible that these results may differ in a PPQ-resistant parasite. In solution, equimolar concentrations of CQ and MFQ inhibited H-FluNox fluorescence to a similar extent, while MFQ had a minimal effect on Ac-H-FluNox fluorescence at even super-pharmacological concentrations. This was somewhat surprising given that MFQ is believed to act in the parasite cytoplasm 102,103. It possible that differences in how MFQ binds to heme 73, affinity of MFQ for other biomolecules 104, and/or modification of MFQ within the parasite could contribute to this discrepancy.

In the early 2000’s, the rise of CQ resistance prompted a search for new combination therapies 105. Several clinical trials were conducted in West Africa to compare efficacy of CQ versus CQ-AS. On day 28, two studies reported similar failure rates for CQ and CQ-AS 106,107. In in a third study, cure rates were significantly higher for CQ-AS compared to CQ alone 108. However, all three studies recommended against use of CQ-AS, due to poor treatment efficacy. Notably, AS-SP and AS-LM were reported to show > 97% efficacy on West African populations during the same time period 109,110. In contrast to superantagonism, in vitro conventional antagonism does not appear to correlate well with treatment failure. ACTs, including DHA-PPQ and AS-ADQ, are highly effective, provided that parasites remain sensitive to one or both ACT component(s). It has yet to be evaluated if antagonistic combinations have silent consequences that might discourage their clinical implementation. For example, since PPQ and ADQ appear to block artemisinin activation, it is possible that these combinations are acting more like pseudo-monotherapies than true combination therapies. This could have a number of downstream consequences, such as impacting the rate of drug resistance evolution. However, future work is needed to address these concerns.

With the absence of new antimalarials ready for clinical implementation, novel strategies with existing drugs have been proposed to overcome current problems with ACT treatment failures. One such strategy, TACTs, has gained the most traction. For traditional ACTs, the short one hour half-life of artemisinins means that long-lived partner drugs are essentially acting as monotherapies for most of the initial 3-day treatment regimen and a substantial period thereafter. In the case of artemisinin-resistant parasites, partner drugs must clear a significantly higher parasite biomass, which increases the likelihood for resistance to emerge. In theory, TACTs are designed to overcome this flaw, whereby two partner drugs with similar half-lives are used, ensuring that parasites are always exposed to at least two antimalarials 111,112. In clinical trials, DHA-PPQ-MFQ showed improved efficacy over DHA-PPQ in regions of high PPQ-resistance 47. However, some have speculated that this was solely due to application of MFQ to MFQ-sensitive parasites and that the TACT would be no more effective than DHA-MFQ 113. PPQ and MFQ were initially promoted as ideal partner drugs for TACTs as it was thought that either due to collateral sensitivity or fitness costs, dual PPQ and MFQ resistance was unlikely to occur 114,115. Use of edited isogenic lines suggests that PPQ resistance conferring mutations in PfCRT have no effect on MFQ sensitivity, while increased MDR1 copy number, which underlies MFQ-resistance, has no effect on PPQ sensitivity 81. Further, more recent genetic surveillance studies of Cambodian parasites cast doubt on the hypothesis that a DHA-PPQ-MFQ resistant parasite would be unfit to spread clinically 116. Previous reports using standard 48 or 72 h assays indicated that PPQ and MFQ are modestly antagonistic 52,55. Interestingly, we observed a much more pronounced superantagonistic phenotype using a pulsing format. It is important to note that unlike artemisinins, PPQ and MFQ have half-lives of several weeks 117,118, raising the question of which assay format would be best suited to evaluate interactions of long-lived partner drugs. Nevertheless, given the interplay of existing PPQ-resistance and PPQ-MFQ antagonism, MFQ and PPQ should, perhaps, be further scrutinized as potential partners for TACTs.

The synthetic ozonide, OZ439, was previously poised to replace artemisinins in combination therapies. While OZ439 is believed to have a mechanism of action similar to that of artemisinins, it has a notably improved 46-62 h half-life 89 and was shown to be effective against artemisinin-resistant parasites 119–121. While OZ439 showed good efficacy alone 89, high treatment failure as part of single-dose combination therapies ultimately resulted in loss of interest of this antimalarial 90–92. However, our data suggest that unsuitable choices in partner drugs may be one of several factors, including poor drug exposure 90–92 and preexisting PPQ-resistance 93,94, that contributed to the poor therapeutic efficacy of these combinations. As assessed by Ac-H-FluNox, both PPQ and FQ appear to rendered heme inert, thus blocking OZ439 activation. Biochemical studies have predicted that due to its less basic behavior, FQ might have a lower propensity to accumulate in the parasite digestive vacuole compared to CQ 95. Greater accumulation of FQ in the parasite cytoplasm compared to other quinolines might explain the ability of FQ to greatly diminish Ac-H-FluNox signal even at nanomolar concentrations. Clinically, peak levels of FQ have been reported to reach approximately 0.5 μM 122, which could significantly impact OZ439 activation. However, unlike CQ, mid-nanomolar FQ concentrations are lethal for parasites, which may explain why we did not observe a super antagonistic phenotype for FQ-OZ439.

The antimalarials that we have are a precious resource, and it is only a matter of time before they fail. Our work clearly demonstrates the importance of detailed understanding of drug-drug interactions at the parasite level. While we believe the pulsing assay used in this study better mimics artemisinin pharmacokinetics compared to standard 72 h assays, it is not suitable for high throughput format. Recently, a high throughput assay was developed to evaluate drug-drug and drug-drug-drug interactions in P. falciparum 55. Using this assay, the authors were able to evaluate interactions of 16 potential TACT combinations against 13 clinical isolates 55. Similar to our findings, the authors reported that interactions can be highly dependent on parasite genetics. This high throughput assay will undoubtably be an important first step in evaluating interactions of new combinations. Nevertheless, since this assay relies on a 72 h format, additive interactions should be interpreted with caution, as this format can mask drug-drug interactions 59. Accordingly, pulsing assays like the one used in this study will be an important step for investigating drug-drug interactions, prior to in vivo studies 123.

Materials and Methods

Reagents

All reagents were purchased from Sigma-Aldrich unless otherwise indicated.

Parasite Culture

Dd2 PfCRTDd2, Dd2PfCRT3D7, Dd2 PfCRTM343L, and Dd2 K13R539T were kindly provided by Dr. David Fidock (Columbia University) 35,64. MRA-1284 was obtained through BEI Resources, NIAID, NIH: Plasmodium falciparum, Strain IPC_6261, MRA-1284, contributed by Didier Ménard. Parasites were propagated in human red blood cells obtained from St. Louis Children’s Hospital blood bank. Cultures were grown at 5% hematocrit in media containing RPMI1640 (Gibco) supplemented with 0.25% (wt/vol) Albumax (Gibco), 15 mg/L hypoxanthine, 110 mg/L sodium pyruvate, 1.19 g/L HEPES, 2.52 g/L sodium bicarbonate, 2 g/L glucose, and 10 mg/L gentamicin and were maintained in air-tight chambers filled with 5% O2/5%CO2/90%N2 at 37°C.

Trophozoite-Stage Dose-Response Assays and Isobolograms

Trophozoite-stage assays were performed as previously described with minor modifications 61. Briefly, parasites were synchronized to the trophozoites stage using two consecutive treatments with 5% sorbitol 10 h apart, followed by culturing for an additional 14 h. Trophozoite stages were seeded at 0.2% parasitemia and 1% hematocrit in round-bottom 96-well plates. Following 4 h incubation with antimalarials, infected red blood cells were washed a minimum of 5 times for plates containing OZ439, and 4 times for all other antimalarials. Plates were centrifuged, 190 μL of media was carefully removed, and 190 μL of fresh media was added. Following the last wash, cultures were transferred to a new flat-bottom 96-well plate. After 72 hr, parasitemia was assessed by flow cytometry. Cultures were resuspended and 10 μL was transferred to a new 96-well plate containing 50 μL per well of 1 x PBS with 1 x SYBR Green I and 100 nM MitoTracker Deep Red. Following incubation at 37°C for 1 h, plates were read on a BD FACSCanto with a BD HTS automated plate reader and analyzed with BD FACSDiva Software version 8. Approximately 10,000 events were analyzed per well. GraphPad version 10 was used to calculate IC50 values by nonlinear regression.

For dose response assays with fixed concentrations of partner drugs, a 96-well plate was prepared containing 2-fold serial dilutions of DHA media at 100 μL per well. A bulk intermediate culture of parasites was prepared at 2% hematocrit. Partner drugs were added to this intermediate culture at 2x concentrations. After mixing, 100 μL of the intermediate culture was added to the 96-well plate containing DHA serial dilutions, for a final hematocrit of 1% and a 1 x fixed partner drug concentration.

For isobolograms, the following fixed ratios were prepared as starting concentrations: 1:0, 4:1, 2:1, 1:1, 1:2, 1:4, 0:1. The starting concentration of each drug alone (1:0 and 0:1 ratios) represents an approximate 16x IC50 concentration. Trophozoite stages were exposed to 2-fold dilutions of these ratios. Fractional IC50 (FIC50) values were calculated as follows: FIC50 values were plotted on an XY graph. Mean ΣFIC50 values were calculated by averaging the sums of FIC50 values from 4:1, 2:1, 1:1, 1:2, and 1:4 ratios. Trophozoite-stage IC50 values for Figure 1-figure supplement 1 and Figure 4-figure supplement 2 were obtained from 1:0 or 0:1 ratios. A one-way ANOVA with a Turkey’s or Dunnett’s multiple comparison test were used to assess statistical significance for IC50 values.

RSAs and Early Ring Stage Dose Response Assays

0-3 hours post invasion (hpi) rings were obtained as previously described 124. Briefly, schizonts were enriched using MACS LD magnet columns (Miltenyi Biotec). Schizont enriched cultures were diluted to 2% hematocrit in 3-4 mL of media and incubated for 3 h with shaking. A 5% sorbitol treatment was then used to remove remaining mature parasites. Early ring stages were seeded at 1% parasitemia and 1% hematocrit in round-bottom 96-well plates. 2-fold serial dilutions were prepared starting at 1400 nM DHA, so that both IC50 values and RSA values could be obtained. After incubation for 3 h, antimalarials were washed out and parasitemia was assessed 72 hr later by flow cytometry as described in the previous section. GraphPad Prism version 10 was used to calculate IC50 values by nonlinear regression. RSA survival was obtained by dividing the parasitemia of cultures treated with 700 nM DHA by the untreated control. A one-way ANOVA with a Dunnett’s multiple comparison test was used to determine statistical significance for IC50 values and RSA survival values.

Detection of Free Heme In Vitro

H-FluNox was kindly provided by Dr. Tasuku Hirayama (Gifu Pharmaceutical University, Lab of Pharmaceutical & Medicinal Chemistry). Reactions were carried out in a buffer containing 50 mM HEPES (pH 7.4) and 100 μM reduced glutathione. Glutathione was used to convert hemin into heme. Hemin was made up in reaction buffer and 2-fold serial dilutions were performed to obtain a range of heme concentrations starting at 500 nM. H-FluNox was used at a final concentration of 500 nM. Reactions were performed in black-walled clear-bottom 96-well plates. Following incubation with H-FluNox and indicated antimalarials at room temperature for 30 min, fluorescence was read at 485/535 nm ex/em using an EnVision Multimode Plate Reader.

Detection of Free Heme In Live Parasites

Ac-H-FluNox was kindly provided by Dr. Tasuku Hirayama (Gifu Pharmaceutical University, Lab of Pharmaceutical & Medicinal Chemistry). Parasites were synchronized to the trophozoite stage as described above. Cultures were then treated with the indicated compound for the indicated time. Media was removed and parasite cultures were resuspended and incubated in 1 x PBS containing 10 μM Ac-H-FluNox for 30 min at 37°C. Ac-H-FluNox is the cell permeable analog of H-FluNox and is converted to H-FluNox in cells 77. Following incubation, cultures were washed with 1x PBS and then imaged using a 63× objective on a Zeiss Imager M2 Plus Widefield fluorescence microscope with AxioVision 4.8 software. Except for Figure 3-figure supplement 1b, a fixed exposure of 350 ms was used to capture H-FluNox fluorescence at 470/509 nm ex/em. Exposure for Figure 3-figure supplement 1b was adjusted for each condition so that DV staining could be optimally visualized. For quantification in Figure 3e,f, 4i and Figure 3-figure supplement 2, at least ten fields were captured per treatment condition and then mean fluorescence intensity of parasite minus background was quantified with ImageJ. Statistical significance of ≥ 70 parasites from at least three independent drug treatments was assessed using a one-way ANOVA.

Western Blot

Parasite drug treatment, protein harvest, sample preparation, SDS-PAGE, and wet transfer were performed as previously described 85. Blots were blocked with 3% BSA in TBS-T and incubated overnight with primary antibodies used at 1:1000 in 1x TBS-T. Anti K48-linked ubiquitin was obtained from Cell Signaling Technologies (catalog number: D9D5, rabbit). Anti-PMV was obtained from 125. Blots were incubated for 1 h at room temperature with IRDye conjugated goat secondary antibodies (LICOR) at 1:10,000 dilutions. Blots were imaged using a Bio Rad imager with Image Lab Touch Software and then quantified with ImageJ. Significant differences between control and antimalarial treated parasites were determined using a one-way ANOVA with a Dunnett’s multiple comparison test.